---
title: "02_SpilloverCompensation_cells_Compressed"
author: "Nathan Steenbuck"
date: "Created: 23 Oct, 2024; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
script_name <- "02_SpilloverCompensation_cells_Uncompressed.Rmd"
cur_user <- Sys.info()[["user"]]

source(file.path("/", "mnt", "central_nas", "projects", 
                 "type1_diabetes", "nathan", "BlockCourse", "2024_BlockCourse", "T1D_analysis", 
                "helpers.R"))
n_cores <- 2
future::plan(future::multicore(workers = n_cores))
paths <- getPaths(script_name)
knitr::opts_knit$set(root_dir = paths$cluster_home)
do_print <- TRUE

# Misc Settings.
seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
knitr::opts_chunk$set(echo = TRUE)
paths$prev <- paste("01_ImportData", paths$object_type, paths$panel_type, sep = "_")
```

You can run this script as: 
Rscript -e "rmarkdown::render('2024_BlockCourse/T1D_analysis/02_SpilloverCompensation_cells_Uncompressed.Rmd')"

# **Goals**

In mass cytometry, a small level of crosstalk between channels can occur. 
Mainly in the channels:
- +1/-1, & +2/-2, due to metal impurities, 
- +16 due to oxidation products.

During data acquisition, this spillover is measured using a spillover 
compensation slide where each spot contains a single metal isotope. 
Here, this spillover measured and correction is applied to the measured cell counts, as 
described in the [original publication](https://www.cell.com/cell-systems/fulltext/S2405-4712(18)30063-2).  
  
This script is based on the [IMC data analysis workflow](https://bodenmillergroup.github.io/IMCDataAnalysis/spillover-correction.html).

# **Settings**  

## Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(SpatialExperiment),
  library(dplyr),
  library(imcRtools),
  library(CATALYST),
  library(ggplot2),
  library(dittoSeq),
  library(patchwork),
  library(pheatmap),
  # library(parallel),
  library(purrr),
  library(tidylog)
))
```


## Paths and settings

Some settings to save plots.

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
seed <- 123456
```


##  Read in the data

Import the SpatialExperiment (SPE) object generated in the first analysis 
script.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```

## Read in the spillover compensation slide
Since, the Hyperion-XTi, IMC data is only outputted as .mcd, not .txt anymore, breaking standard pipelines.
Therefore, we have converted the .mcds files to .tiffs files (without HPF) using the `steinbock` package.
Now, we continue with the .tiffs.

```{r get-directories}
## Directory with :tiffs
directory <- file.path(paths$folder_in, "img_spillcomp", paths$panel_type)
if (!dir.exists(directory)) stop("Directory with .tiffs not found")

# Image metadata
fn_image_metadata <- file.path(paths$folder_in, paste0("images_", paths$panel_type, "_spillcomp.csv"))
images <- readr::read_csv(fn_image_metadata, show_col_types = FALSE)

if (!file.exists(fn_image_metadata)) stop("Image metadata file not found")

# Panel information.
panel <- rowData(spe)
```


Check for .tiff files in the directory.
Loop through files, and write the pixel intensities to a dataframe.
```{r read-tiffs}
files <- list.files(directory, pattern = "\\.tiff$", full.names = TRUE)
total <- data.frame()

# Loop through files in the directory
for (file_name in files) {
  # Read TIFF image
  numeric_part <- basename(file_name)
  img <- EBImage::readImage(file_name)
  
  # Process the image (assuming similar processing as in your code)
  img <- lapply(1:nrow(panel), function(x) as.vector(img[,,x]))
  #img <- lapply(1:nrow(panel), function(x) as.vector(img[,,x]))
  img <- as.data.frame(img)
  colnames(img) <- 1:length(colnames(img))
  img$spot <- rep(numeric_part, nrow(img))
  
  # Bind the data to the total dataframe
  total <- rbind(total, img)
}
```

31x mass channels open. Not one missing?
Convert to tibble. Rename with metal names.
```{r}
# nr of pixels
nrow(total)

# Nr. of unique spots = 37. Including ROI_001.
unique(total$spot)

# Nr. of channels
ncol(total) - 1
```


```{r path-spillcomp}
## The names were read-in incorrectly, so we need to fix them. 
total <- total |> as_tibble() |> 
  rename_with(~ panel$metal, -spot) |> 
  mutate(spot = ifelse(spot == "SpillComp_Uncompressed_Batch1_ROI_001_1_ROI_001.tiff", "ROI_001", spot)) |>
  mutate(spot = gsub("SpillComp_Uncompressed_Batch1_", "", spot)) |> 
  # Ho_165_33_Ho_165.tiff into Ho16533 -> to troubleshoot wrong naming.
  mutate(spot_id = gsub("([A-Za-z]+)_([0-9]+)_([0-9]+).*", "\\1\\2\\3", spot)) |> 
  # mutate(spot_id = gsub("([A-Za-z]+)_([0-9]+).*", "\\1\\2", spot)) |>
  select(-spot)

# 36 unique spots (without ROI_001).
length(unique(total$spot_id))
unique(total$spot_id)
```

Always identify for a given spot, the column with the highest value and assign it to the corresponding metal.
```{r}
## Identifying strongest metal.
results <- total |> 
  group_by(spot_id) |>
  summarise(across(1:(nrow(panel)-1), \(.x) mean(.x, na.rm = TRUE))) |> 
  tidyr::pivot_longer(cols = 2:(nrow(panel)), names_to = "metal", values_to = "intensity") |> 
  group_by(spot_id) |>
  summarise(metal = metal[which.max(intensity)])

# Dictionary for translation.
results <- results |> 
  filter(spot_id != "ROI_001") |>
  # Bi209 counts very low.
  mutate(metal = ifelse(spot_id == "Dy16213", "Ho165", metal)) |> 
  mutate(metal = ifelse(spot_id == "Nd14315", "Bi209", metal)) |>
  # IMPORTANT: Remove Compressed channels: 
  filter(!spot_id %in% c("Ho1655", "Yb17136", "Sm1494", "Sm1523", "Nd1436"))

## Join corrected metal names.
total <- total |> 
  inner_join(results, by = "spot_id") |> 
  # Now we know which metal 
  select(-spot_id)
```

Now, read in as SCE.
```{r read-spillover-sce}
all_counts <- total[,-(nrow(panel)+1)]
sce <- SingleCellExperiment(assays = list(counts = t(all_counts)))

colData(sce) <- DataFrame(total[,(nrow(panel)+1), drop=FALSE])
rowData(sce)$sample_mass <- as.numeric(gsub("\\D", "", rownames(sce)))
colData(sce)$sample_mass <- as.numeric(gsub("\\D", "", sce$metal))
rowData(sce)$channel_name <- paste0(rownames(sce), "Di")
assay(sce, "exprs") <- asinh(counts(sce)/5)
```


## Spillover heatmaps

Next, the median pixel intensities per spot are plotted and thresholded on 
medians higher than 200 counts. Low median intensities (< 200 counts) can 
hinder robust spillover estimation, in which case consecutive pixels can be 
binned.  
  
The median pixel intensity per channel should be highest for the corresponding 
metal. If this is not the case, either the .TXT file name is incorrect or the 
wrong metal isotope was spotted on the spillover compensation slide.  


```{r spillcomp-heatmaps, fig.height=7, fig.width=7}
imcRtools::plotSpotHeatmap(sce, spot_id = "metal", statistic = "mean", cluster_rows = FALSE, cluster_cols= FALSE, 
  order_metals = TRUE)

plotSpotHeatmap(sce, spot_id = "metal", log = FALSE, threshold = 100)
```

## Pixel binning

Consecutive pixels are summed to increase the robustness of the spillover 
estimation. No signal should be observed off the diagonal on the thresholded 
heatmap. If that is the case, reduce the bin size.

```{r binning, fig.height=7, fig.width=7}
# Pixel binning
bin_size <- 20

sce <- binAcrossPixels(sce, spot_id = "metal", bin_size = bin_size)
assay(sce, "exprs") <- asinh(counts(sce)/5)

## Plot binned heatmaps.
plotSpotHeatmap(sce, spot_id = "metal", statistic = "mean", cluster_rows = FALSE, cluster_cols= FALSE, 
  order_metals = TRUE, assay_type = "counts")

## Binned + Thresholded heatmaps:
plotSpotHeatmap(sce, spot_id = "metal", log = FALSE, threshold = 200, 
  cluster_rows = FALSE, cluster_cols= FALSE, order_metals = TRUE, assay_type = "counts")
```

## Filter pixels

### Debarcoding and pixel assignment

Pixels are assigned to their corresponding "barcode" (in this case the already 
known metal isotope spot). This allows to identify pixels that cannot be 
robustly assigned to the spotted metal. These background pixels are removed 
prior to spillover estimation:  
1. Assign a preliminary metal mass to each pixel.  
2. For each pixel, estimate a cutoff parameter for the distance between 
positive and negative pixel sets.  
3. Apply the estimated cutoffs to identify truly positive pixels.  

```{r}
# Get correct key.
bc_key <- as.numeric(unique(sce$sample_mass))
bc_key <- bc_key[order(bc_key)]

sce <- assignPrelim(sce, bc_key = bc_key)
sce <- estCutoffs(sce)
sce <- applyCutoffs(sce)

cur_table <- table(sce$bc_id, sce$sample_mass)
cur_table
```

```{r}
library(pheatmap)
cur_table <- table(sce$bc_id, sce$sample_mass)

# Visualize the correctly and incorrectly assigned pixels
pheatmap(log10(cur_table + 1),
         cluster_rows = FALSE,
         cluster_cols = FALSE)
        
# Compute the fraction of unassigned pixels per spot
cur_table["0",] / colSums(cur_table)
```


# **Compute spillover matrix**

## Compute and plot spillover matrix

Functions from the [CATALYST package](https://bioconductor.org/packages/release/bioc/html/CATALYST.html) are
used to compute the spillover matrix and visualize it. 

```{r compute-spillover, fig.width=7, fig.height=7}
sce <- computeSpillmat(sce)

isotope_list <- CATALYST::isotope_list
isotope_list$Ar <- 80

plotSpillmat(sce, isotope_list = isotope_list)
```

Save Spillover matrix in the metadata of the SCE object + as external csv.
```{r}
# Save spillover matrix in variable
sm <- metadata(sce)$spillover_matrix

write.csv(sm, "sm.csv")
```

# **Compute spillover matrix**

## Compute and plot spillover matrix

Functions from the [CATALYST package](https://bioconductor.org/packages/release/bioc/html/CATALYST.html) are
used to compute the spillover matrix and visualize it. 

```{r apply-spillover-compensation}
# Apply spillover compensation per batch. This takes considerable time.
rowData(spe)$channel_name <- paste0(rowData(spe)$metal, "Di")
spillmat <- metadata(sce)$spillover_matrix
metadata(spe)$spillover_matrix <- spillmat

spe <- CATALYST::compCytof(spe, spillmat, transform = FALSE, cofactor = 1,
                            isotope_list = isotope_list, overwrite = FALSE)

# Extract compensated counts
comp_counts <- assay(spe, "compcounts")

# Overwrite the compensated counts
assay(spe, "raw") <- assay(spe, "counts")
assay(spe, "counts") <- comp_counts

# Write to exprs:
assay(spe, "exprs_pre") <- asinh(assay(spe, "raw")/5)
assay(spe, "exprs") <- asinh(assay(spe, "counts")/5)
```

```{r}
library(dittoSeq)
library(patchwork)
before <- dittoScatterPlot(spe, x.var = "CD4", y.var = "CD31",
                           assay.x = "exprs_pre", assay.y = "exprs") +
    ggtitle("Before compensation")

after <- dittoScatterPlot(spe, x.var = "CD4", y.var = "CD31",
                          assay.x = "exprs", assay.y = "exprs") +
    ggtitle("After compensation")
before + after
```

# **Save the updated SPE object**

Save the updated SPE object containing the spillover-compensated counts (as 
`counts`). Original (uncompensated) counts are stored as "raW" and will be 
deleted in the next script.

```{r save-spe}
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
print(spe)
saveRDS(spe, fn_spe)
```
