---
title: "03_TransformCorrect_cells_Compressed"
author: "Nathan Steenbuck"
date: "Created: 23 Oct, 2024; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
script_name <- "03_TransformCorrect_cells_Compressed.Rmd"

source(file.path("/", "mnt", "central_nas", "projects", 
                 "type1_diabetes", "nathan", "BlockCourse", "2024_BlockCourse", "T1D_analysis", 
                "helpers.R"))
n_cores <- 2
future::plan(future::multicore(workers = n_cores))
paths <- getPaths(script_name)
knitr::opts_knit$set(root_dir = paths$cluster_home)
do_print <- TRUE

# Misc. Settings.
seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
options(future.globals.maxSize = 40 * 1024 * 1024 * 1024)
knitr::opts_chunk$set(echo = TRUE)
paths$prev <- paste("02_SpilloverCompensation", paths$object_type, paths$panel_type, sep = "_")
```


The script can be run directly from the command line with:
Rscript -e "rmarkdown::render('2024_BlockCourse/T1D_analysis/03_TransformCorrect_cells_Compressed.Rmd')"

# **Goals**

**1. Counts transformations**  
IMC data can be transformed to achieve more symmetric intensity distributions with more similar ranges across markers.  

# **Settings**  

## Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(SpatialExperiment),
  library(dplyr),
  library(batchelor),
  library(BiocParallel),
  library(ggplot2),
  library(scater)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Import the SpatialExperiment (SPE) object generated in the first analysis script.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```



## Remove uncompensated counts

Remove `raw` (uncompensated) counts (not needed anymore).

```{r remove-raw-counts}
if ("raw" %in% assayNames(spe)) {
  assay(spe, "raw") <- NULL
}
```



# **Data transformation**

IMC data can be transformed to achieve more symmetric intensity distributions with more similar ranges across markers. 
In line with CyTOF data, IMC data is often arcsinh-transformed. 
However, the choice of the so called "cofactor" has to be adjusted for IMC data due its lower average intensity. 
IMC data are often arcsinh-transformed using a cofactor of 1. 

## Arcsinh transformation

Here, we will arsinh transform the data using a cofactor of 1.  
Other transormation methods, such as log transformation can also be used. 
For visualization purpose, arcsinh-transformed data can be further scaled using z-score standardization.

### Cell counts

```{r asinh-transf}
# Arcsinh
if (!"exprs" %in% assayNames(spe))
  assay(spe, "exprs") <- asinh(counts(spe) / 1)
```

## Quantile normalization

Alternatively, the raw data can be scaled between 0 and 1 after clipping outlying intensities to the 99.9th percentile value.

### Cell counts

```{r scaled-transf}
# Define the censor value
censor_val <- 0.999

# Calculate quantile
if (!"scaled" %in% assayNames(spe)) {
  quant <- apply(assay(spe, "counts"), 1, quantile, probs = censor_val)
  
  # Normalize
  assay(spe, "scaled") <- apply(assay(spe, "counts"), 2, function(x) x / quant)
  
  # Clip values 0-1
  assay(spe, "scaled")[assay(spe, "scaled") > 1] <- 1
  assay(spe, "scaled")[assay(spe, "scaled") < 0] <- 0
}
```

## Save the updated SPE object

Contains the new counts transformations.

```{r save-sce1}
print(assayNames(spe))
fn_spe <- file.path(paths$folder_out, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
saveRDS(spe, fn_spe)
```
