---
title: "03_TransformCorrect_cells_Compressed"
author: "Nathan Steenbuck"
date: "Created: 23 Oct, 2024; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
script_name <- "03_TransformCorrect_cells_Compressed.Rmd"

source(file.path("/", "mnt", "central_nas", "projects", 
                 "type1_diabetes", "nathan", "BlockCourse", "2024_BlockCourse", "T1D_analysis", 
                "helpers.R"))
n_cores <- 2
future::plan(future::multicore(workers = n_cores))
paths <- getPaths(script_name)
knitr::opts_knit$set(root_dir = paths$cluster_home)
do_print <- TRUE

# Misc. Settings.
seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
options(future.globals.maxSize = 40 * 1024 * 1024 * 1024)
knitr::opts_chunk$set(echo = TRUE)
paths$prev <- paste("02_SpilloverCompensation", paths$object_type, paths$panel_type, sep = "_")
```


The script can be run directly from the command line with:
Rscript -e "rmarkdown::render('2024_BlockCourse/T1D_analysis/03_TransformCorrect_cells_Compressed.Rmd')"

# **Goals**

**1. Counts transformations**  
IMC data can be transformed to achieve more symmetric intensity distributions with more similar ranges across markers.  

# **Settings**  

## Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(SpatialExperiment),
  library(dplyr),
  library(batchelor),
  library(BiocParallel),
  library(ggplot2),
  library(scater)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Import the SpatialExperiment (SPE) object generated in the first analysis script.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```



## Remove uncompensated counts

Remove `raw` (uncompensated) counts (not needed anymore).

```{r remove-raw-counts}
if ("raw" %in% assayNames(spe)) {
  assay(spe, "raw") <- NULL
}
```



# **Data transformation**

IMC data can be transformed to achieve more symmetric intensity distributions with more similar ranges across markers. 
In line with CyTOF data, IMC data is often arcsinh-transformed. 
However, the choice of the so called "cofactor" has to be adjusted for IMC data due its lower average intensity. 
IMC data are often arcsinh-transformed using a cofactor of 1. 

## Arcsinh transformation

Here, we will arsinh transform the data using a cofactor of 1.  
Other transormation methods, such as log transformation can also be used. 
For visualization purpose, arcsinh-transformed data can be further scaled using z-score standardization.

### Cell counts

```{r asinh-transf}
# Arcsinh
if (!"exprs" %in% assayNames(spe))
  assay(spe, "exprs") <- asinh(counts(spe) / 1)
```

## Quantile normalization

Alternatively, the raw data can be scaled between 0 and 1 after clipping outlying intensities to the 99.9th percentile value.

### Cell counts

```{r scaled-transf}
# Define the censor value
censor_val <- 0.999

# Calculate quantile
if (!"scaled" %in% assayNames(spe)) {
  quant <- apply(assay(spe, "counts"), 1, quantile, probs = censor_val)
  
  # Normalize
  assay(spe, "scaled") <- apply(assay(spe, "counts"), 2, function(x) x / quant)
  
  # Clip values 0-1
  assay(spe, "scaled")[assay(spe, "scaled") > 1] <- 1
  assay(spe, "scaled")[assay(spe, "scaled") < 0] <- 0
}
```


# Batch Effect correction.

## Parameters

Define the channels to correct and the subset of cells to use to calculate the batch effect (see `?"batchelor-restrict"`).
This is used later in fastMNN.
The H3 and DNA channels are not used, as we do not expect cell type specific variance for these channels.
This is used as a simple feature selection step.

```{r fastMNN-parameters}
# Define the subset of cells per donor to use to calculate the batch effect -> nb_cells * 95 
nb_cells <- 1000

set.seed(seed)
cell_restrict <- as.data.frame(cbind(rn = rownames(colData(spe)),
                                     case_id = colData(spe)$case_id)) |> 
  group_by(case_id, .add = TRUE) |>
  sample_n(nb_cells) |>
  pull(rn)

# Subset the SPE to the channels to correct.
spe_sub <- spe[, cell_restrict]
```

## **FastMNN**

Batch correction is performed with [FastMNN](http://bioconductor.org/packages/release/bioc/vignettes/batchelor/inst/doc/correction.html#3_mutual_nearest_neighbors) 
from the [batchelor package](https://bioconductor.org/packages/devel/bioc/html/batchelor.html).  

### Run FastMNN
For now, we subset SPE to 1000 cells per donor. These are thought to be representative of the donor.

Repeat for Case_id and Batch.
Works better for Case_id variable.

```{r fastMNN-run-caseid, message=FALSE}
do_fast_mnn <- TRUE
method_name <- "fastMNN"
batch_variables <- c("case_id") # , "batch")
channels_to_correct <- !(grepl("DNA|H3", rownames(spe)))

## Run FastMNN for both batch variables.
results <- furrr::future_map(batch_variables, \(batch_variable) {
  assay_name <- paste0(method_name, "_", batch_variable)
  message(assay_name)
  if (!assay_name %in% assayNames(spe)) {
    do_fast_mnn <- TRUE
    set.seed(seed)
    # Correct by donor
    spe_mnn <- fastMNN(spe,
                       batch = spe[[batch_variable]],
                       restrict = list(cell_restrict),
                       subset.row = channels_to_correct,
                       assay.type = "exprs",
                       d = 40,
                       auto.merge = TRUE,
                       correct.all = TRUE,
                       cos.norm = FALSE,
                       BPPARAM = MulticoreParam(),
                       BSPARAM = BiocSingular::ExactParam())
    # Extract metdata related to batch correction
    metadata_merge <- metadata(spe_mnn)$merge.info
    list("spe_mnn" = spe_mnn, "metadata" = metadata_merge)
  }
}, .options = options) |> purrr::set_names(batch_variables)

## Add the corrected counts to the SPE.
assay_name1 <- paste0(method_name, "_", batch_variables[[1]])
assay(spe,  assay_name1) <- as.matrix(assay(results[[1]][["spe_mnn"]], "reconstructed"))
metadata_merges <- list("Donor" = results[[1]][["metadata"]]) 
```


## Save the updated SPE object

Contains the new counts transformations.

```{r save-sce1}
print(assayNames(spe))
fn_spe <- file.path(paths$folder_out, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
saveRDS(spe, fn_spe)
```
