---
title: "04_QualityControl_cells_Uncompressed"
author: "Nathan Steenbuck, Nicolas Damond"
date: "Created: 23 May, 2022; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
script_name <- "04_QualityControl_cells_Uncompressed.Rmd"

source(file.path("/", "mnt", "central_nas", "projects", 
                 "type1_diabetes", "nathan", "BlockCourse", "2024_BlockCourse", "T1D_analysis", 
                "helpers.R"))
n_cores <- 2
future::plan(future::multicore(workers = n_cores))
paths <- getPaths(script_name)
knitr::opts_knit$set(root_dir = paths$cluster_home)
do_print <- TRUE

# Misc. Settings.
seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
options(future.globals.maxSize = 50 * 1024 * 1024 * 1024)
knitr::opts_chunk$set(echo = TRUE)
paths$prev <- paste("03_TransformCorrect", paths$object_type, paths$panel_type, sep = "_")
```

## **Goals**  
  
Perform quality controls on cell-level data.   
  
**Diagnostic plots**  
- Number of cells and images per condition  
- Density plots  
- 90th percentile per channel
- Heatmaps

**Dimensionality reduction**  
- UMAP  
- Plot key variables (stage, cases) on reduced dimensions.  
- Plot marker abundances on reduced dimensions.  

**Image visualization**  
- Display images with `cytoviewer` for visual quality check.  
- Display images before/after spillover-compensation.  



# **Settings**  

## Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(parallel),
  library(furrr),
  library(tictoc)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```



# **Diagnostic plots**

```{r packages2, results='hide', include=FALSE}
suppressPackageStartupMessages(c(
  library(vroom),
  library(RColorBrewer),
  library(ggplot2),
  library(ggridges),
  library(scuttle),
  library(scater),
  library(dittoSeq),
  library(heatmaply),
  library(htmltools),
  library(BiocParallel),
  library(ggrepel),
  library(EBImage)
  #library(foreach),
  #library(doParallel)
))
```

## Order rows

Load a file that indicates in which order the data should be plotted.
The content of this file is then saved in the metadata slot of the SPE object.

```{r row-order}
# Load the file
fn_order <- file.path(paths$folder_in, "order.csv")
if (!file.exists(fn_order)) stop("Ordering file not found")
row_order <- vroom(fn_order, show_col_types = FALSE)

# Row order for cases and stages
stages <- row_order[!is.na(row_order$stages), ]$stages
stages <- stages[stages %in% spe$donor_type]
cases  <- row_order[!is.na(row_order$cases), ]$cases
cases <- cases[cases %in% spe$case_id]

# Save to metadata
metadata(spe)[["stages"]] <- stages
metadata(spe)[["cases"]] <- cases
```

## Select assays and channels

Modify the `assay_sel` variable to select which assay (for instance scaled counts (`scaled`) or asinh-transformed counts (`exprs`)) should be used for quality control plots. 
Multiple assays can be entered but this leads to increased processing times.

```{r select-assays}
# Select the assay(s) of interest
assay_sel <- c("counts", "scaled", "exprs")
writeLines(c("Assays:", assay_sel[assay_sel %in% assayNames(spe)]))

# Select the channels to display
channels <- rownames(spe)[!(grepl("DNA|H3", rownames(spe)))]
writeLines(c("\nNumber of channels:", length(channels)))

# Variables to plot
plot_variables <- c("stages", "cases")
plot_variables2 <- c("stages", "casestages")
names(plot_variables2) <- names(plot_variables) <- c("donor_type", "case_id")
writeLines("\nVariables to plot:")
print(plot_variables)
```

Change the case_id column to a factor.
```{r factor_column}
colData(spe)[, "case_id"] <- factor(colData(spe)[, "case_id"])
```

## Subset the SPE object

The SPE object is subset to `nb_cells` per case for faster plotting. Hence, for example `2000 cells` per donor.

```{r subset-spe}
if (!("subset" %in% names(metadata(spe)))) {
  # Cells per case
  nb_cells <- 7500
  
  # Subset the SPE object (nb_cells per case)
  cell_subset <- tibble(rn = rownames(colData(spe)),
                        case_id = colData(spe)$case_id) |> 
    group_by(case_id) |>
    sample_n(nb_cells) |> 
    pull(rn)
  
  # Keep the subset cell ids in SPE metadata
  metadata(spe)[["subset"]] <- sort(as.vector(cell_subset))
}

# Keep only the selected cells and channels
spe_sub <- spe[channels, metadata(spe)[["subset"]]]
```


## Number of images and cells per case and stage

Plot the number of images per case and stage.

```{r images-per-condition}
# Calculate number of images per case and stage
cur_dat <- as_tibble(colData(spe)) |>
  dplyr::distinct(case_id, donor_type, image_id) |>
  dplyr::count(case_id, donor_type) |>
  dplyr::mutate(case_id = factor(case_id, levels = metadata(spe)$cases),
                donor_type = factor(donor_type, levels = metadata(spe)$stages)) |>
  dplyr::arrange(case_id, donor_type)

# Plot number of images per case
p <- cur_dat |>
  ggplot(aes(y = n,
             x = as.factor(donor_type),
             color = as.factor(donor_type))) +
  geom_boxplot(outlier.shape = NA) +
  geom_point()+
  expand_limits(y = 0) +
  scale_color_manual(values = palettes$stages) +
  labs(color = "T1D stage", x = NULL, y = "Number of images per case") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)

fn <- paste0(paste(today, "Number", "ImagesPerCase", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# Plot total number of images per disease stage
p <- cur_dat |>
  ggplot(aes(y = n,
             x = as.factor(donor_type),
             fill = as.factor(donor_type))) +
  geom_bar(stat = "identity") +
  expand_limits(y = 0) +
  scale_fill_manual(values = palettes$stages) +
  labs(fill = "T1D stage", x = NULL, y = "Total number of images per stage") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)

fn <- paste0(paste(today, "Number", "ImagesPerStage", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

Plot the number of measured cells per case and per stage.

```{r cells-per-condition}
# Calculate number of cells per case and stage
cur_dat <- as_tibble(colData(spe)) |>
  group_by(case_id, donor_type) |>
  summarise(n = n(), .groups = "keep") |>
  mutate(case_id = factor(case_id, levels = metadata(spe)$cases),
         donor_type = factor(donor_type, levels = metadata(spe)$stages)) |>
  arrange(case_id, donor_type)

# Plot number of cells per case
p <- cur_dat |>
  ggplot(aes(y = n,
             x = as.factor(donor_type),
             color = as.factor(donor_type))) +
  geom_boxplot(outlier.shape = NA) +
  geom_point()+
  expand_limits(y = 0) +
  scale_color_manual(values = palettes$stages) +
  labs(color = "T1D stage", x = NULL, y = "Number of cells per case") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p

fn <- paste0(paste(today, "Number", "CellsPerCase", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# Plot number of cells per disease stage
p <- cur_dat |>
  ggplot(aes(y = n,
             x = as.factor(donor_type),
             fill = as.factor(donor_type))) +
  geom_bar(stat = "identity") +
  expand_limits(y = 0) +
  scale_fill_manual(values = palettes$stages) +
  labs(fill = "T1D stage", x = NULL, y = "Total number of cells per stage") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p

fn <- paste0(paste(today, "Number", "CellsPerStage", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

Plot the number of cells per image.

```{r cells-per-image}
# Calculate number of cells per case and stage
cur_dat <- as_tibble(colData(spe)) |>
  group_by(case_id, donor_type, image_id) |>
  summarise(n = n(), .groups = "keep") |>
  mutate(case_id = factor(case_id, levels = metadata(spe)$cases),
         donor_type = factor(donor_type, levels = metadata(spe)$stages)) |>
  arrange(case_id, donor_type)

# Plot number of cells per image and per case
p <- cur_dat |>
  ggplot(aes(y = n,
             x = as.factor(case_id),
             color = as.factor(donor_type))) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(size = 0.5) +
  expand_limits(y = 0) +
  scale_color_manual(values = palettes$stages) +
  labs(color = "Case", x = NULL, y = "Number of cells per image") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
p

fn <- paste0(paste(today, "Number", "CellsPerImage", "byCase",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))


# Plot number of cells per image and per stage
cur_dat <- cur_dat |>
  group_by(case_id, donor_type) |>
  summarise(avg = mean(n), .groups = "keep")
  
p <- cur_dat |>
  ggplot(aes(y = avg,
             x = as.factor(donor_type),
             color = as.factor(donor_type))) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  expand_limits(y = 0) +
  scale_color_manual(values = palettes$stages) +
  labs(color = "T1D stage", x = NULL, y = "Average number of cells per image") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p

fn <- paste0(paste(today, "Number", "CellsPerImage", "byStage",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```


## Density plots

For the different transformations, check the distribution of each marker intensity to identify potential global shifts and drop outs.

-> That is, compare densities between donors (and stages) for each channel.

```{r density-plots, message=FALSE, fig.height=10, fig.width=15}
cur_assay <- "exprs"

# Create tibble with image number, cell number, case and intensity values.
cur_dat <- dplyr::bind_cols(
  image_numer = colData(spe_sub)$image_number,
  cell_number = colData(spe_sub)$cell_number,
  case_id = colData(spe_sub)$case_id,
  t(assay(spe_sub, cur_assay)))

# Pivot longer into channel and intensity columns
cur_dat <- cur_dat |>
  tidyr::pivot_longer(cols = -c(image_numer, cell_number, case_id), 
                      names_to = "channel", values_to = cur_assay)  |> 
  dplyr::filter(channel %in% channels)  |> 
  dplyr::arrange(case_id)

# Plot Intensity distribution for all channels and donors.
p <- cur_dat |> 
  ggplot(aes(x = get(cur_assay), y = as.factor(case_id))) +
  geom_density_ridges(aes(fill = as.factor(case_id), color = "black"),
                      alpha = 0.5) +
  facet_wrap(~channel) +
  scale_fill_manual(values = palettes$casestages) +
  scale_color_manual(values = palettes$stages) +
  ggtitle(paste("ChannelsperDonor", "-", cur_assay)) +
  guides(fill = FALSE) +
  labs(x = cur_assay, y = "Case", color = "T1D stage") +
  mytheme$standard() +
  theme(axis.text.y = element_text(size = 3))
p  
# Save.
fn <- paste0(paste(today, "DensityPlots", "ChannelsperDonor",
                    cur_assay, sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

## Heatmaps

### Visualize mean expression levels

Heatmaps are generated with the [dittoSeq package](http://www.bioconductor.org/packages/release/bioc/html/dittoSeq.html).  

Prepare the data:  
* Calculate mean marker intensity per case.
* Transform counts.

```{r calculate-mean-expr}
# Workaround for *** hardcoded `sample_id` in `SpatialExperiment`
sce <- as(spe, "SingleCellExperiment")
sce$sample_id <- sce$image_id

# Calculate mean intensities
sce_mean <- scuttle::aggregateAcrossCells(sce,
                                          ids = sce$case_id,
                                          statistics = "mean",
                                          use.dimred = FALSE)
colnames(sce_mean) <- sce_mean$case_id
remove(sce)

# Asinh-transformed counts
assay(sce_mean, "exprs") <- asinh(assay(sce_mean, "counts"))

# Quantile-normalized counts
censor_val <- 0.999
quant <- apply(assay(sce_mean, "counts"), 1, quantile, probs = censor_val)
assay(sce_mean, "scaled") <- apply(assay(sce_mean, "counts"), 2,
                                   function(x) x / quant)
assay(sce_mean)[assay(sce_mean, "scaled") > 1] <- 1
assay(sce_mean, "scaled")[assay(sce_mean, "scaled") < 0] <- 0

# Select assays for ploting mean marker intensities
assay_sel_mean <- c("counts", "exprs", "scaled")
```

Heatmaps of marker intensities without scaling

```{r dittoheatmap-not-scaled}
cur_assay <- "scaled"
h <- dittoHeatmap(sce_mean, assay = cur_assay,
                  annot.by = c("case_id", "donor_type"),
                  order.by = match(cases, sce_mean$case_id),
                  scale = "none", heatmap.colors = viridis(100),
                  cluster_cols = FALSE,
                  main = paste("Heatmap", "NotScaled", cur_assay,
                                sep = " - "),
                  annotation_colors = list(case_id = palettes$cases,
                                            donor_type = palettes$stages),
                  genes = channels)
fn <- paste0(paste(today, "HeatmapDitto", "NotScaled",
                    cur_assay, sep = "_"), ".png")
do.call(ggsave, c(list(fn, h), plotsave_param))
```

# Images
# **Image Visualization**

The [cytomapper](https://github.com/BodenmillerGroup/cytomapper/) package can be used to visualize IMC data. 
For this, the names and locations of the image files must be retrieved and added to the spe object.

```{r packages4, results='hide'}
suppressPackageStartupMessages(c(
  library(cytomapper), library(cytoviewer)
))
```


## Parameters

The following files are required: 
- Image stacks (multichannel `.tiff` stacks)  
- Cell masks (`tiff` images)  

A few random images (number defined in `nb_images`) are loaded and displayed.   

**Folder names and file name patterns must be changed manually**  

```{r image-settings}
# Number of images to load and display
nb_images <- 14
image_extension <- ".tiff"
segmentation_type <- "whole-cell"

# Folders
folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
if (!dir.exists(folder_images)) stop("Image folder not found")
folder_masks <- file.path(paths$folder_in, "masks_cells",
                          paths$panel_type, segmentation_type)
if (!dir.exists(folder_masks)) stop("Mask folder not found")

# Select random images and the associated masks
set.seed(seed)
image_list <- list.files(folder_images)
image_list <- mask_list <- sort(sample(unique(image_list), nb_images))
writeLines(c("Selected images:", image_list))
```


## Load images

Load full image stacks and cell masks as `CytoImageList` objects.

```{r load-images}
# Load multichannel images
images <- imgloader(
  x = spe,
  image_dir = folder_images,
  image_names = image_list,
  type = "stacks"
)

# Load cell masks
masks <- imgloader(
  x = spe,
  image_dir = folder_masks,
  image_names = mask_list,
  as.is = TRUE,
  # bit.depth = ,
  type = "masks"
)

if (length(setdiff(mcols(images)$ImageName,
                   mcols(masks)$ImageName)) != 0) {
  print("mcols(images)$ImageName and mcols(masks)$ImageName
        should have the same values")
}

# Subset the SPE object
spe_sub <- spe[, spe$image_fullname %in% image_list]
spe_sub$ImageName <- gsub(image_extension, "", spe_sub$image_fullname)
```

Inspect the images.
```{r display-images}
library(cytoviewer)
app <- cytoviewer(image = images, 
                  mask = masks, 
                  object = spe_sub,
                  cell_id = "cell_number", 
                  img_id = "ImageName")

if (interactive()) {
    shiny::runApp(app)
}
```

Q: What can you observe in the images?
Q: Do you see the previously identified differences in marker expression between the donors?  

## Image-level quality control

Image-level quality control is often performed using tools that offer a graphical user interface such as QuPath, FIJI and the previously mentioned cytoviewer package. Viewers that were specifically developed for IMC data can be seen here. In this section, we will specifically focus on quantitative metrics to assess image quality.

It is often of interest to calculate the signal-to-noise ratio (SNR) for individual channels and markers. Here, we define the SNR as:

*SNR=Is/In*

where *Is* is the intensity of the signal (mean intensity of pixels with true signal) 
and *In* is the intensity of the noise (mean intensity of pixels containing noise). 
This definition of the SNR is just one of many and other measures can be applied. 
Finding a threshold that separates pixels containing signal and pixels containing noise is not trivial and different approaches can be chosen. Here, we use the otsu thresholding approach to find pixels of the “foreground” (i.e., signal) and “background” (i.e., noise). The SNR is then defined as the mean intensity of foreground pixels divided by the mean intensity of background pixels. We compute this measure as well as the mean signal intensity per image. 
The plot below shows the average SNR versus the average signal intensity across all images.

Note 1: The SNR is calculated on the pixel level and not on the cell level.
Note 2: The fitting can fail on the image-level, which produces Infinity values.
```{r snr-images}
library(tidyverse)
library(ggrepel)
library(EBImage)

cur_snr <- lapply(names(images), function(x){
    img <- images[[x]]
    mat <- apply(img, 3, function(ch){
        # Otsu threshold
        thres <- otsu(ch, range = c(min(ch), max(ch)), levels = 65536)
        # Signal-to-noise ratio
        snr <- mean(ch[ch > thres]) / mean(ch[ch <= thres])
        # Signal intensity
        ps <- mean(ch[ch > thres])
        
        return(c(snr = snr, ps = ps))
    })
    t(mat) %>% as.data.frame() %>% 
        mutate(image = x,
               marker = colnames(mat)) %>% 
        pivot_longer(cols = c(snr, ps))
})

cur_snr <- do.call(rbind, cur_snr)

p <- cur_snr %>% 
    group_by(marker, name) %>%
    summarize(log_mean = log2(mean(value))) %>%
    pivot_wider(names_from = name, values_from = log_mean) %>%
    ggplot() +
    geom_point(aes(ps, snr)) +
    geom_label_repel(aes(ps, snr, label = marker)) +
    mytheme$standard()+  
    ylab("Signal-to-noise ratio [log2]") +
    xlab("Signal intensity [log2]")

fn <- paste0(paste(today, "SNRvsSignalIntensity", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

## Cell-level quality control

In the following paragraphs we will look at different metrics and visualization approaches to assess data quality (as well as biological differences) on the single-cell level.

Related to the signal-to-noise ratio (SNR) calculated above on the pixel-level, a similar measure can be derived on the single-cell level. Here, we will use a two component Gaussian mixture model for each marker to find cells with positive and negative expression. The SNR is defined as:

**SNR=Is/In**

where *Is* is the intensity of the signal (mean intensity of cells with positive signal) 
and *In* is the intensity of the noise (mean intensity of cells lacking expression). 
To define cells with positive and negative marker expression, we fit the mixture model across the transformed counts of all cells contained in the SpatialExperiment object. Next, for each marker we calculate the mean of the non-transformed counts for the positive and the negative cells. 
The SNR is then the ratio between the mean of the positive signal and the mean of the negative signal.

```{r snr-cells}
library(mclust)

set.seed(220224)
mat <- sapply(seq_len(nrow(spe)), function(x){
    cur_exprs <- assay(spe, "exprs")[x,]
    cur_counts <- assay(spe, "counts")[x,]
    
    cur_model <- Mclust(cur_exprs, G = 2)
    mean1 <- mean(cur_counts[cur_model$classification == 1])
    mean2 <- mean(cur_counts[cur_model$classification == 2])
    
    signal <- ifelse(mean1 > mean2, mean1, mean2)
    noise <- ifelse(mean1 > mean2, mean2, mean1)
    
    return(c(snr = signal/noise, ps = signal))
})
    
cur_snr <- t(mat) %>% as.data.frame() %>% 
        mutate(marker = rownames(spe))

p <- cur_snr %>% ggplot() +
    geom_point(aes(log2(ps), log2(snr))) +
    geom_label_repel(aes(log2(ps), log2(snr), label = marker)) +
    mytheme$standard() + 
    ylab("Signal-to-noise ratio [log2]") +
    xlab("Signal intensity [log2]")
print(p)
fn <- paste0(paste(today, "SNRvsSignalIntensityCell", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

# **Dimensionality reduction**

Dimensionality reduction algorithms are applied on different count transformations (`assays`) and stored in the `reducedDims` of the SPE object. 
In addition, selected variables and marker expression are plotted on the obtained reduced dimensions.  

Dimensionality reduction algorithms currently used:  
- **UMAP** [uwot package](https://github.com/jlmelville/uwot).


## Packages

```{r packages3, results='hide'}
suppressPackageStartupMessages(c(
  library(uwot),
  library(viridis),
  library(RSpectra)
))
```

## Select parameters

- Dimensionality reduction algorithm
- Assays (count transformations)
- Channels
- Variables to plot

```{r select-assay2}
# Select the dimensionality reduction algorithms to use
dimred_sel <- c("UMAP") 
writeLines(c("Dimensionality reduction:", dimred_sel))

# Select the assay(s) of interest
assay_sel <- c("scaled", "exprs")
writeLines(c("\nAssays:", assay_sel[assay_sel %in% assayNames(spe)]))

# Select the channels
channels_dimred <- rowData(spe)[rowData(spe)$dimred == 1, ]$shortname
cat(c("Channels:", channels_dimred[channels_dimred %in% rownames(spe)]))
cat(c("\nNumber of channels:", length(channels_dimred)))

# Variables to plot
plot_variables <- c("stages", "cases")
names(plot_variables) <- c("donor_type", "case_id")
writeLines("\nVariables to plot:")
print(plot_variables)

# Subset the SPE object
spe_sub <- spe[channels_dimred, metadata(spe)[["subset"]]]
cells_not_subset <- colnames(spe)[! colnames(spe) %in% metadata(spe)$subset]
```


## Run dimensionality reduction

### Run UMAP

UMAP for every selected assay, only if there is no object with the same name in `reducedDimNames(spe)`.  
UMAP is performed all cells.

If this takes too long: the results can also be generated by a subset, and then be extrapolated to all cells - 
see publication.

```{r run-UMAP}
# Save as variable and write assays to NULL. Reduces memory overhead for parallelization. 
colnames_spe <- colnames(spe)

# Add UMAPs to SPE object
cur_assay <- "exprs"
dimred_name <- paste("UMAP", cur_assay, sep = "_")
print(dimred_name)

# Run UMAP on a cell subset
if ((!dimred_name %in% reducedDimNames(spe_sub)) && ("UMAP" %in% dimred_sel)) {
  # Extract Counts. 
  if (cur_assay %in% assayNames(spe_sub)) {
    counts <- t(assay(spe_sub, cur_assay))
  }
  # Run UMAP.
  umap_model <- uwot::umap(counts, ret_model = TRUE)
  # Extract Embedding.
  cur_umap <- umap_model$embedding
  colnames(cur_umap) <- c("UMAP1", "UMAP2")
  rownames(cur_umap) <- rownames(counts)
}

reducedDim(spe_sub, dimred_name) <- cur_umap
```

## Plot reduced dimensions

### Plot variables

Plot selected variables on reduced dimensions.

```{r plot-dimred, fig.width=15, fig.height=10}
# Prepare the data
cur_dat <- makePerCellDF(spe_sub) |>
  dplyr::mutate(case_id = factor(case_id, levels = cases),
                donor_type = factor(donor_type, levels = stages)) |>
  dplyr::arrange(case_id, donor_type)

cur_dimred <- "UMAP"
# Plot variables
purrr::iwalk(plot_variables, \(cur_var, name_cur_var) {
  dimred_name <- paste(cur_dimred, cur_assay, sep = "_")
  message(dimred_name)
  p <- plot_dim_red(cur_dat, dimred_name, name_cur_var,
                    size = 0.5, palette = palettes[[cur_var]])
  
  fn <- paste0(paste(today, dimred_name, cur_var, sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
})
```

Questions:
Do you see differences on the UMAPs?
If so, what does that mean?

### Plot marker expression

```{r plot-dimred-channels, fig.height=10, fig.width=15}
# Plot marker expression
dimred_name <- paste(cur_dimred, cur_assay, sep = "_")

# Prepare the data
if (cur_assay %in% assayNames(spe_sub)) {
  cur_dat <- scuttle::makePerCellDF(spe_sub, features = channels,
                                    exprs_values = cur_assay) |>
              dplyr::select(c(all_of(channels), paste0(dimred_name, ".1"), 
                                                paste0(dimred_name, ".2")))
} else {
  cur_dat <- scuttle::makePerCellDF(spe_sub, features = channels,
                                    exprs_values = "exprs") |>
    dplyr::select(c(all_of(channels), paste0(dimred_name, ".1"), 
                                      paste0(dimred_name, ".2")))
}

cur_dat <- cur_dat |>
  tidyr::pivot_longer(cols = all_of(channels),
                      names_to = "channel",
                      values_to = cur_assay)

# Plot marker expression
p <- plot_dim_red_channels(cur_dat, dimred_name, cur_assay,
                            channels, force_points = TRUE)

fn <- paste0(paste(today, dimred_name, "Channels", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```