---
title: "05_CellCategories_cells_Compressed"
author: "Nathan Steenbuck"
date: "Created: 23 Oct, 2024; Compiled: `r format(Sys.time(), '%d %b, %')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
script_name <- "05_CellCategories_cells_Compressed.Rmd"

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("04_QualityControl", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```



# **Goal**

In the following scripts, cell types are attributed to all cells in the dataset
in an iterative way:

- `05_CellCategories_cells_Compressed` (this script): Attribution of main 
categories (immune, islet, other (exocrine+stroma)).
- `06_CellTypes_cells_Compressed`: Attribution of cell types.

In both scripts this is performed by performing PhenoGraph clustering using the Rphenoannoy package.

The resulting cell categories, which are used in downstream analyses are stored
as `colData(spe)$cell_category`.


# **Settings**

## Load packages

```{r packages, results="hide"}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(parallel),
  library(tictoc),
  library(purrr),
  library(furrr)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```



# **Clustering**

## Settings

### Load packages

RPhenoannoy can installed with: `devtools::install_github("stuchly/Rphenoannoy@8b81e2e7fb0599f45070e2cba1b28ac219b7c472")`

```{r packages2, results="hide"}
suppressPackageStartupMessages(c(
  library(ggplot2),
  library(scater),
  library(scuttle),
  library(scran),
  library(igraph),
  library(Rphenoannoy),
  library(clustree),
  library(ranger),
  library(patchwork),
  library(BiocParallel),
  library(tictoc)
))
```

### Channels and assays

Select channels (`channels_clust`), assays (`assay_sel`) and clustering methods (`methods_sel`) to use.

```{r select-methods-assays}
methods_sel <- c("Pheno1")

assay_sel <- c("scaled")
names(assay_sel) <- c("scaled")

writeLines(c("Assays:", assay_sel[assay_sel %in% assayNames(spe)], 
                        assay_sel[assay_sel %in% reducedDimNames(spe)]))

dimred_sel <- c("UMAP")
writeLines(c("\nReduced dimensions:", dimred_sel))

channels <- rownames(spe)[!(grepl("DNA|H3", rownames(spe)))]
cat(c("\nChannels:", channels[channels %in% rownames(spe)]))
cat(c("\nNumber of channels:", length(channels)))
```

### Select channels to use for clustering

Reduced dimension plots showing marker expression that generated by the previous script 
can be used to select the most relevant markers for clustering.

```{r select-channels-clustering}
channels_clust <- rownames(rowData(spe)[rowData(spe)$clustering == 1, ])
cat(c("\nChannels used for unsupervised clustering:",
      channels_clust[channels_clust %in% rownames(spe)]))
```


## Subset

```{r convert-to-sce}
# Extract the metadata (to avoid conflicts when merging the SCEs)
meta <- metadata(spe)
spatial_coords <- spatialCoords(spe)
colpairs <- colPairs(spe)

# Convert to SingleCellExperiment
sce <- as(spe, "SingleCellExperiment")
remove(spe)
```

## **Clustering: PhenoGraph**

Unsupervised clustering is performed with the [PhenoGraph](https://doi.org/10.1016/j.cell.2015.05.047) algorithm. 
This method works by generating a nearest-neighbor (kNN) graph of phenotpyic similarities followed by Louvain community dectection. 

Here, we try the [Rphenoannoy implementation](https://github.com/stuchly/Rphenoannoy).
RPhenoannoy implements a parallel Jaccard-coefficient, approximates the kNN and uses Louvain clustering.

Other implementations such as FastPG were tested, but were eventually dropped to its non-reproducibility.

```{r phenograph-run}
clust_method <- c("Pheno1")
cur_assay <- assay_sel
# Number of nearest-neighbors
k <- 30

## 
clust_name <- paste(clust_method, cur_assay, sep = "_")
writeLines(c("\n", clust_name))

# Run Phenograph
if (!clust_name %in% colnames(colData(sce))) {
  set.seed(seed)
  ## Try Rphenoannoy. 950 k cells = 17 min!
  cur_pheno_annoy <- Rphenoannoy::Rphenoannoy(t(assay(sce, cur_assay)), k = k)

  cur_pheno <- DataFrame(cur_pheno_annoy[[2]]$membership)
  colnames(cur_pheno) <- clust_name
  rownames(cur_pheno) <- colnames(assay(sce, cur_assay))

  # Add Phenograph clusters to the colData of the SCE object
  # Cluster `0` is attributed to non-subsetted cells and not islet cells
  colData(sce)[, clust_name] <<- cur_pheno
  remove(cur_pheno)
}
```



### Save SCE
```{r save-sce}
FIXME: adjust here.
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
message(assayNames(sce))
message(reducedDimNames(sce))
saveRDS(sce, fn_spe)
```



# **Visualize clusters**

## Load packages

```{r packages3, results="hide"}
suppressPackageStartupMessages(c(
  library(heatmaply),
  library(htmltools)#,
  #library(cytomapper)
))
```


## Visualize clusters on reduced dimensions

```{r dimred-clusters}
cur_dat <- makePerCellDF(sce_sub, use_dimred = TRUE) |>
  dplyr::mutate(case_id = factor(case_id, levels = meta$cases),
                donor_type = factor(donor_type, levels = meta$stages)) |>
  dplyr::arrange(case_id, donor_type)  |> 
  tibble::as_tibble()

purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    purrr::walk(methods_sel, \(cur_method) {
      dimred_name <- paste(cur_dimred, cur_assay, sep = "_")
      clust_name <- paste(cur_method, cur_assay, sep = "_")

      # Plot all clustersdat, dimred, color_by
      p <- plot_dim_red(dat = cur_dat, dimred = dimred_name, color_by = clust_name,
                        sample = TRUE, size = 0.1, alpha = 1)
      if (do_print) print(p)
      fn <- paste0(paste(today, "Clusters", clust_name, cur_dimred,
                         sep = "_"), ".png")
      do.call(ggsave, c(list(fn, p), plotsave_param))
    })
  })
})
```


## Distribution of marker expression by cluster

```{r violin-clusters-markers, message=FALSE, warning=FALSE, fig.width=15, fig.height=10}
purrr::map(assay_sel, \(cur_assay) {
  purrr::map(methods_sel, \(cur_method) {
    # cur_assay <- assay_sel[2]; cur_method <- methods_sel[1]
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    print(clust_name)
    sce_sub[[clust_name]] <- as.factor(sce_sub[[clust_name]])

    if (cur_assay %in% SummarizedExperiment::assayNames(sce_sub)) {
        cur_dat <- scuttle::makePerCellDF(sce_sub, assay.type = cur_assay,
                                          features = channels_clust, use.dimred = FALSE)
    }

    cur_dat <- melt.data.table(
      as.data.table(cur_dat),
      id.vars = c("image_number", "cell_number", clust_name),
      measure.vars = channels_clust,
      variable.name = "channel",
      value.name = cur_assay
    )
    cur_dat[, MeanExprs := mean(get(cur_assay)),
            by = c("channel", clust_name)]

    p <- plot_violin(cur_dat, x = "channel", y = cur_assay,
                     fill_by = "MeanExprs", facet_by = clust_name,
                     title = clust_name) +
      theme(axis.text.x = element_text(size = 8, angle = 90))
    if (do_print) suppressMessages(print(p))
    fn <- paste0(paste(today, "Clusters", clust_name, "Violins",
                       sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
  })
})
```


## Heatmap of marker expression by cluster
```{r heatmap-clusters-markers}
hmlist <- list()
n <- 1
names(assay_sel) <- c("counts", "scaled")
purrr::walk(methods_sel, \(cur_method) {
  purrr::walk2(assay_sel, names(assay_sel), \(cur_assay, name_cur_assay) {
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    message(clust_name)
    message(name_cur_assay)
    # Summarize the data

    hm <- summarize_heatmap(sce_sub,
                            expr_values = name_cur_assay,
                            cluster_by = clust_name,
                            channels = channels_clust)
    
    # Display the heatmap
    fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap",
                       sep = "_"), ".html")
    hmlist[[n]] <- heatmaply(
       heatmaply::normalize(hm), main = clust_name, 
       file = file.path(paths$folder_script, fn))
    n <- n + 1
    
    # remove all clusters with less than 10 cells and plot again.
    clust_freq <- table(colData(sce_sub)[[clust_name]])
    keep_clusts <- names(clust_freq)[clust_freq >= 10]
    sce_isl_sub <- sce_sub[, colData(sce_sub)[[clust_name]] %in% keep_clusts]

    hm <- summarize_heatmap(sce_isl_sub,
                            expr_values = name_cur_assay,
                            cluster_by = clust_name,
                            channels = channels_clust)
    # Display the reduced heatmap -> this improves the visualization.
    # Otherwise heatmap is scaled by expression of outliers.
    fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap_reduced",
                       sep = "_"), ".html")
    hmlist[[n]] <- heatmaply(
       heatmaply::normalize(hm), main = clust_name, 
       file = file.path(paths$folder_script, fn))
    n <- n + 1

  })
})
htmltools::tagList(setNames(hmlist, NULL))
names(assay_sel) <- c("exprs", "scaled")
```



## Visualize clusters with cytomapper

### Select cluster(s) and assay to show

```{r viz-clust-select}
viz_clust <- NULL # Select cluster(s) to visualize.
viz_method <- "Pheno1"
viz_assay <- "scaled"
```

### Load images and masks

```{r viz-cluster-load}
if (!is.null(viz_clust)) {
  nb_images <- 14
  image_extension <- ".tiff"
  clust_name <- paste(viz_method, viz_assay, sep = "_")

  # Subset the SCE
  sce_viz <- sce[, colData(sce)[[clust_name]] %in% viz_clust]

  # Select random image.
  set.seed(seed)
  image_sub <- sort(sample(
    unique(sce_viz$image_fullname),
  min(length(unique(sce_viz$image_fullname)), nb_images)))

  # Folders
  folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
  folder_masks <- file.path(paths$folder_in, "masks_cells",
                            paths$panel_type, "whole-cell")

  # Load images and masks
  images <- yoloader(
    x = sce_viz,
    image_dir = folder_images,
    image_names = image_sub,
    type = "stacks"
  )

  masks <- yoloader(
    x = sce_viz,
    image_dir = folder_masks,
    image_names = image_sub,
    as.is = TRUE,
    type = "masks"
  )

  sce_viz <- sce_viz[, sce_viz$image_fullname %in% image_sub]
  sce_viz$ImageName <- gsub(image_extension, "", sce_viz$image_fullname)
}
```

### Interactive with Cytoviewer

```{r viz-cluster-cytoviewer}
# Use cytoviewer with images, masks and object
# library(cytoviewer)
if (!is.null(viz_clust)) {
  channels_view <- channels_clust
  sub_images <- cytomapper::getChannels(images, channels_view)
  app <- cytoviewer(image = sub_images, 
                    mask = masks, 
                    object = sce_viz[channels_view, ], 
                    img_id = "ImageName", 
                    cell_id = "cell_number")

  if (interactive()) {
    shiny::runApp(app)
  }
}
```



# **Attribute cell categories**

**Note: this section requires manual intervention**

Here, clusters obtained at the previous steps are manually merged into meaningful cell categories.
Cluster numbers are attributed to the different cell types based on the plots and heatmaps above. 
Consequently, this attribution has to be adapted to the clustering results.

## Cell category attribution

Cell category attribution has to be performed for every combination of clustering methods and selected assays.

### Phenograph x scaled counts

Depending on the clustering, maybe select other categories.

```{r cellcats-pheno-scaled}
clust_methods <- c("Pheno1")
clust_assay <- c("scaled")

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")
if (length(clust_assay) != 1)
  stop("Select only one assay")

clust_name <- paste(clust_methods, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
clust_immune <- c()
clust_islet  <- c()
clust_ductal <- c()
clust_acinar <- c()
clust_exocrine <- c()
clust_fibro_endo <- c()
clust_other <- c()

all_clust <- sort(c(clust_immune, clust_islet, clust_ductal, clust_acinar,
                    clust_exocrine, clust_fibro_endo, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
colData(sce)[colData(sce)[, clust_name] %in% clust_immune,
             paste("CellCat", clust_name, sep = "_")] <- "Immune"
colData(sce)[colData(sce)[, clust_name] %in% clust_islet,
             paste("CellCat", clust_name, sep = "_")] <- "Islet"
colData(sce)[colData(sce)[, clust_name] %in% clust_ductal,
             paste("CellCat", clust_name, sep = "_")] <- "Ductal"
colData(sce)[colData(sce)[, clust_name] %in% clust_acinar,
             paste("CellCat", clust_name, sep = "_")] <- "Acinar"
colData(sce)[colData(sce)[, clust_name] %in% clust_exocrine,
             paste("CellCat", clust_name, sep = "_")] <- "Exocrine"
colData(sce)[colData(sce)[, clust_name] %in% clust_fibro_endo,
             paste("CellCat", clust_name, sep = "_")] <- "Fibro_Endo"
colData(sce)[colData(sce)[, clust_name] %in% clust_other,
             paste("CellCat", clust_name, sep = "_")] <- "Other"
```



## Visualize attributed categories

```{r assay-sel}
assay_sel <- c("scaled")
names(assay_sel) <- c("scaled")
```

### Plot cell types on reduced dimensions


```{r cellcats-dimred}
# Prepare the data
cur_dat <- makePerCellDF(sce[, metadata(sce)$subset], use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = metadata(sce)$cases),
         donor_type = factor(donor_type, levels = metadata(sce)$stages)) |>
  arrange(case_id, donor_type)

# Plot
purrr::map(assay_sel, \(cur_assay) {
  purrr::map(dimred_sel, \(cur_dimred) {
    purrr::map(methods_sel, \(cur_method) {
      dimred_name <- paste(cur_dimred, cur_assay, sep = "_")
      clust_name <- paste("CellCat", cur_method, cur_assay,
                          sep = "_")

      p <- plot_dim_red(cur_dat, dimred_name, clust_name,
                        sample = TRUE, size = 0.1, alpha = 1)
      if (do_print) { print(p)}

      fn <- paste0(paste(today, clust_name, cur_dimred,
                         sep = "_"), ".png")
      do.call(ggsave, c(list(fn, p), plotsave_param))
    })
  })
})
```

### Cell categories heatmap

```{r cellcats-heatmap}
hmlist <- list()
n <- 1
purrr::map2(assay_sel, names(assay_sel), \(cur_assay, name_cur_assay) {
  purrr::map(methods_sel, \(cur_method) {
    clust_name <- paste("CellCat", cur_method, cur_assay, sep = "_")

    # Summarize the data
    hm <- summarize_heatmap(sce[, meta$subset],
                            expr_values = name_cur_assay,
                            cluster_by = clust_name,
                            channels = channels_clust)

    # Display the heatmap
    fn <- paste0(paste(today, clust_name, "Heatmap",
                       sep = "_"), ".html")

    hmlist[[n]] <- heatmaply(
      heatmaply::normalize(hm), main = clust_name,
      file = file.path(paths$folder_script, fn))
    n <- n + 1
  })
})
htmltools::tagList(setNames(hmlist, NULL))
```

### Visualize categories with cytomapper

Load the masks

```{r viz-cellcats-load}
nb_images <- 14
image_extension <- ".tiff"
# clust_name <- paste(viz_method, viz_assay, sep = "_")

set.seed(seed)
image_sub <- sort(sample(
  unique(sce$image_fullname),
  min(length(unique(sce$image_fullname)), nb_images)))

# Subset the SCE to subsampled images
sce_sub <- sce[, sce$image_fullname %in% image_sub]
sce_sub$ImageName <- gsub(image_extension, "", sce_sub$image_fullname)

# Folders
folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
folder_masks <- file.path(paths$folder_in, "masks_cells",
                          paths$panel_type, "whole-cell")

# Load masks
masks <- yoloader(
  x = sce_sub,
  image_dir = folder_masks,
  image_names = image_sub,
  as.is = TRUE,
  type = "masks"
)
```

Plot the images

```{r viz-cellcats-plot}
purrr::map(assay_sel, \(cur_assay) {
  purrr::map(methods_sel, \(cur_method) {
    celltype_viz <- paste("CellCat", cur_method, cur_assay,
                          sep = "_")

    colour_key <- list(c(
      Immune = palettes$colors[1], Islet = palettes$colors[8],
      Ductal = palettes$colors[5], Acinar = palettes$colors[4],
      Exocrine = palettes$colors[3], Fibro_Endo = palettes$colors[11],
      Other = palettes$colors[13]))
    names(colour_key) <- celltype_viz

    fn <- list(filename = file.path(
      plotsave_param$path,
      paste0(paste(today, "Cytomapper", celltype_viz, sep = "_"), ".png")))

    cytomapper::plotCells(masks, object = sce_sub,
              cell_id = "cell_number", img_id = "ImageName",
              colour_by = celltype_viz, colour = colour_key, save_plot = fn)
  })
})
```



# **Consensus cell categories**

Attribute consensus cell categories based on the clustering performed in the previous script.

## Load packages

```{r packages4, results="hide"}
suppressPackageStartupMessages(c(
  library(ggalluvial)
))
```


## Attribute temporary cell types

Exocrine cells are grouped into a single category and the ductal/acinar types are stored into temporary cell types.
For the other cells, the attributed cell type corresponds to the cell category. 
Precise cell types will be refined in downstream scripts.

```{r attribute-celltypes}
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    celltype_name <- paste("CellType", clust_name, sep = "_")
    cellcat_name <- paste("CellCat", clust_name, sep = "_")

    colData(sce)[, celltype_name] <<- colData(sce)[, cellcat_name]

    colData(sce)[colData(sce)[, cellcat_name] %in% c(
      "Exocrine", "Acinar", "Ductal"), cellcat_name] <<- "exocrine"
    colData(sce)[colData(sce)[, cellcat_name] == "Immune",
                 cellcat_name] <<- "immune"
    colData(sce)[colData(sce)[, cellcat_name] == "Islet",
                 cellcat_name] <<- "islet"
    colData(sce)[colData(sce)[, cellcat_name] == "Fibro_Endo",
                 cellcat_name] <<- "mesenchymal_endothelial"
    colData(sce)[colData(sce)[, cellcat_name] == "Other",
                 cellcat_name] <<- "other"
  })
})
```


## Cell category concordance

Compare concordance of cell categories attributed using different methods.

```{r cellcats-concordance}
cellcat_cols <- colnames(colData(sce))[grepl(paste0(
  "CellCat_", methods_sel, collapse = "|"), colnames(colData(sce)))]

cell_categories <- as.data.table(colData(sce)[
  , c("cell_id", "case_id", "image_number", "cell_number", cellcat_cols)])

ccat_combination <- t(combn(unique(cellcat_cols), 2))

for (i in seq_len(nrow(ccat_combination))) {
  cat("\n", ccat_combination[i, 1], " vs ", ccat_combination[i, 2], "\n")

  ccat_concord <- nrow(cell_categories[
    get(ccat_combination[i, 1]) == get(ccat_combination[i, 2]), ]
  )
  ccat_discord <- nrow(cell_categories[
    get(ccat_combination[i, 1]) != get(ccat_combination[i, 2]), ]
  )
  cat("Number of cells with concordant cell categories: ", ccat_concord,
      "\nNumber of cells with discordant cell categories: ", ccat_discord,
      "\nPercentage of cells with concordant cell categories:",
      100 * ccat_concord / (ccat_discord + ccat_concord), "%", "\n")
}
```
Pheno1-scaled vs Leiden1-fastMNN: 79% (88%)
Pheno1-scaled vs Pheno1-fastMNN: 74% (83%)
Pheno1-scaled vs Leiden1-scaled: 91.2 % (92.6%)
Leiden1-fastMNN vs Pheno1-fastMNN: 77 % (88%)
Pheno1-fastMNN vs Leiden1-scaled: 72 % (82%)
Leiden1-fastMNN vs Leiden1-scaled: 74 % (85 %)

## Attribute consensus cell categories

### Majority vote

Here, we remove Pheno1-fastMNN. 
Cells for which 2 out of 3 methods agree on the cell categories are attributed this category, other cells as considered as "ambiguous".

```{r consensus-cell-categories}
# Remove a column matching "Pheno1_fastMNN"
# Use grep to find the column that matches the pattern "Pheno1-fastMNN"
column_to_remove <- grep("Pheno1_fastMNN", names(cell_categories))

# Remove the column by setting it to NULL
cell_categories[, (column_to_remove) := NULL]

# Majority vote
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

cell_categories[, cell_category := apply(cell_categories, 1, Mode)]
cell_categories[, Paste := do.call(paste0, .SD),
                .SDcols = grep("CellCat_", colnames(cell_categories))]
cell_categories[, count := stringr::str_count(Paste, cell_category)]
cell_categories[count < 2, cell_category := "ambiguous"]
cell_categories[, `:=`(Paste = NULL, count = NULL)]

cat("Percentage of cells with an ambiguous cell type: ", 100 * nrow(
  cell_categories[cell_category == "ambiguous"]) / nrow(cell_categories), "%")
```

### Compare cell type attributions

```{r cellcat-alluvial, warning=FALSE}
cellcat_cols <- c("cell_category", colnames(colData(sce))[grepl(paste0(
  "CellCat_", methods_sel, collapse = "|"), colnames(colData(sce)))])

cellcat_cols <- cellcat_cols[-grep("Pheno1_fastMNN", cellcat_cols)]

p <- cell_categories[, .(number = .N), by = cellcat_cols] |>
  ggplot(aes(axis1 = get(cellcat_cols[1]),
             axis2 = get(cellcat_cols[2]),
             axis3 = get(cellcat_cols[3]),
             axis4 = get(cellcat_cols[4]),
             # axis5 = get(cellcat_cols[5]),
             y = number)) +
  geom_stratum() +
  geom_text(stat = "stratum", infer.label = TRUE) +
  geom_alluvium(aes(fill = get(cellcat_cols[1]))) +
  scale_x_discrete(limits = gsub("cell_category_", "", cellcat_cols)) +
  scale_fill_manual(values = palettes$colors[c(TRUE, FALSE)],
                    name = "Cell category") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
if (do_print) { print(p)}

fn <- paste0(paste(today, "CellCat", "alluvial", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

### Number of cells by cell category

```{r cells-per-category}
as_tibble(cell_categories) |>
  select(cell_category) |>
  group_by(cell_category) |>
  add_count(name = "Number of cells") |>
  distinct()
```




REMOVE UNNEEDED COLUMNS:
CellType_Pheno1_scaled / CellType_Leiden1_scaled 
CellType_Pheno1_fastMNN / CellType_Leiden1_fastMNN
Pheno1_fastMNN / Leiden1_fastMNN / Pheno1_scaled / Leiden1_scaled 




## Add celltypes to the main SCE

```{r add-celltypes-sce}
cell_categories <- cell_categories[order(match(cell_categories$cell_id,
                                               colnames(sce))), ]
colData(sce)$cell_category <- cell_categories$cell_category
```



# **Plot cell categories**

## Reduced dimensions

Plot consensus cell categories on reduced dimensions

```{r consensus-dimred}
ccat <- c("immune", "islet", "exocrine", "mesenchymal_endothelial",
          "ambiguous", "other")
names(ccat) <- c("Immune", "Islet", "Exocrine", "Mesenchymal_Endothelial",
                 "Ambiguous", "Other")
palette_ccat <- c(palettes$colors[1:(length(ccat) - 2)], "grey40", "grey")
names(palette_ccat) <- ccat


cur_dat <- scuttle::makePerCellDF(sce[, colnames(sce) %in% metadata(sce)$subset],
                                  use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = metadata(sce)$cases),
         donor_type = factor(donor_type, levels = metadata(sce)$stages)) |>
  arrange(case_id, donor_type)

purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    dimred_name <- paste(cur_dimred, cur_assay, sep = "_")

    p <- plot_dim_red(cur_dat, dimred_name, "cell_category",
                      sample = TRUE, size = 0.1, alpha = 1,
                      palette = palette_ccat)
    if (do_print) print(p)

    fn <- paste0(paste(today, "CellCat", "Consensus", dimred_name,
                       sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
  })
})
```

## Heatmap

```{r consensus-heatmap}
hmlist <- list()

purrr::map(assay_sel, \(cur_assay) {
  clust_name <- paste("cell_category", cur_assay, sep = "_")

  # Summarize the data
  hm <- summarize_heatmap(sce,
                          expr_values = cur_assay,
                          cluster_by = "cell_category",
                          channels = channels)

  # Display the heatmap
  fn <- paste0(paste(today, "CellCat", "Consensus", cur_assay,
                     "Heatmap", sep = "_"), ".html")

  hmlist[[i]] <- heatmaply(
    heatmaply::normalize(hm), main = clust_name,
    file = file.path(paths$folder_script, fn))
})
htmltools::tagList(setNames(hmlist, NULL))
```


## Correlation between channels

```{r cell-category-correlation}
# Define the correlation method
cor_method <- "pearson"

# Function to calculate p values
cor_test_p <- function(x) {
  FUN <- function(x, y) cor.test(x, y, method = cor_method)[["p.value"]]
  z <- outer(
    colnames(x),
    colnames(x),
    Vectorize(function(i, j) FUN(x[, i], x[, j]))
  )
  dimnames(z) <- list(colnames(x), colnames(x))
  z
}

hmlist <- list()
purrr::map2(assay_sel, names(assay_sel), \(cur_assay, name_cur_assay) {

  # Summarize the data
  hm <- summarize_heatmap(sce,
                          expr_values = name_cur_assay,
                          cluster_by = "cell_category",
                          channels = channels)

  # Run the correlation test
  r <- cor(hm, method = cor_method)
  p <- cor_test_p(hm)
  diag(p) <- NA

  # Plot and save the heatmap
  fn <- paste0(paste(today, "CellCat", "Consensus", "Correlation",
                     cur_assay, sep = "_"), ".html")

  hmlist[[i]] <- heatmaply::heatmaply_cor(
    r, node_type = "scatter",
    point_size_mat = -log10(p), point_size_name = "-log10(p-value)",
    label_names = c("x", "y", "Correlation"),
    main = paste("Cell type -", cur_assay),
    file = file.path(paths$folder_script, fn))
})
htmltools::tagList(setNames(hmlist, NULL))
```

## Images

Load random images and masks

```{r consensus-viz-load}
nb_images <- 14
image_extension <- ".tiff"

# Folders
folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
folder_masks <- file.path(paths$folder_in, "masks_cells",
                          paths$panel_type, "whole-cell")

# Select random images and the associated masks
set.seed(seed)
image_list <- list.files(folder_images)
image_list <- mask_list <- sort(sample(unique(image_list), nb_images))
writeLines(c("Selected images:", image_list))

# Load images and masks
images <- yoloader(
  x = sce,
  image_dir = folder_images,
  image_names = image_list,
  type = "stacks"
)

masks <- yoloader(
  x = sce,
  image_dir = folder_masks,
  image_names = mask_list,
  as.is = TRUE,
  type = "masks"
)

if (length(setdiff(mcols(images)$ImageName,
                   mcols(masks)$ImageName)) != 0) {
  print("mcols(images)$ImageName and mcols(masks)$ImageName
        should have the same values")
}

# Subset the SCE object
sce_viz <- sce[, sce$image_fullname %in% image_list]
sce_viz$ImageName <- gsub(image_extension, "", sce_viz$image_fullname)
```

Plot the images

```{r viz-celltype-plot}
fn <- list(filename = file.path(
  plotsave_param$path,
  paste0(paste(today, "CellCat", "Cytomapper", sep = "_"), ".png")))

color_key <- list(c(
  immune = "red", islet = "blue",  exocrine = "yellow",
  mesenchymal_endothelial = "green",
  ambiguous = "white", other = "grey30"))
names(color_key) <- "cell_category"

image_names <- c(unique(colData(sce_viz)[, c("ImageName", "case_id")])$case_id)

cytomapper::plotPixels(images, object = sce_viz, masks,
           cell_id = "cell_number", img_id = "ImageName",
           colour_by = c("SYP", "CD163", "MPO", "CD3e", "PDX1", "CAV1"),
           outline_by = "cell_category",
           # image_title = list(text = image_names),
           colour = color_key,
           bcg = list(SYP = c(0, 5, 1), CD163 = c(0, 5, 1), MPO = c(0, 7, 1),
                      CD3e = c(0, 5, 1), PDX1 = c(0, 3, 1), CAV1 = c(0, 4, 1)),
           image_title = NULL, scale = TRUE, save_plot = fn)
```



# **Save **

## Remove temporary columns

```{r remove-cellcat-columns}
cellcat_cols <- colnames(colData(sce))[grep("CellCat_", colnames(colData(sce)))]
colData(sce)[, cellcat_cols] <- NULL
```

## Convert to SPE
