---
title: "05_CellCategories_cells_Uncompressed"
author: "Nathan Steenbuck"
date: "Created: 23 Oct, 2024; Compiled: `r format(Sys.time(), '%d %b, %')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
script_name <- "05_CellCategories_cells_Compressed.Rmd"

source(file.path("/", "mnt", "central_nas", "projects", 
                 "type1_diabetes", "nathan", "BlockCourse", "2024_BlockCourse", "T1D_analysis", 
                "helpers.R"))
n_cores <- 2
future::plan(future::multicore(workers = n_cores))
paths <- getPaths(script_name)
knitr::opts_knit$set(root_dir = paths$cluster_home)
do_print <- TRUE

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("04_QualityControl", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```



# **Goal**

In the following scripts, cell types are attributed to all cells in the dataset
in an iterative way:

- `05_CellCategories_cells_Compressed` (this script): 
  - Attribution of cell categories from the compressed dataset
- `06_CellCategories_cells_Compressed`:
  - Attribution of cell types.

This is performed by performing PhenoGraph clustering using the Rphenoannoy package.

For cell type annotation we perform unsupervised clustering, and then relate the lineage marker expression of clusters to cell types.

The resulting cell categories, which are used in downstream analyses are stored
as `colData(spe)$cell_category`.

Here, the idea is to use the MINIMAL dataset to generate cell categories that can be used for the T1D dataset.

# **Settings**

## Load packages

```{r packages, results="hide"}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(parallel),
  library(purrr),
  library(furrr)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```

# **Clustering**

## Settings

### Load packages

RPhenoannoy can installed with: `devtools::install_github("stuchly/Rphenoannoy@8b81e2e7fb0599f45070e2cba1b28ac219b7c472")`

```{r packages2, results="hide"}
suppressPackageStartupMessages(c(
  library(ggplot2),
  library(scater),
  library(scuttle),
  library(scran),
  library(Rphenoannoy),
  library(patchwork),
  library(BiocParallel)
))
```

### Channels and assays

Select channels (`channels_clust`), assays (`assay_sel`) and clustering methods (`methods_sel`) to use.

```{r select-methods-assays}
methods_sel <- c("Pheno1")

assay_sel <- c("scaled")
names(assay_sel) <- c("scaled")

writeLines(c("Assays:", assay_sel[assay_sel %in% assayNames(spe)], 
                        assay_sel[assay_sel %in% reducedDimNames(spe)]))

dimred_sel <- c("UMAP")
writeLines(c("\nReduced dimensions:", dimred_sel))

channels <- rownames(spe)[!(grepl("DNA|H3", rownames(spe)))]
cat(c("\nChannels:", channels[channels %in% rownames(spe)]))
cat(c("\nNumber of channels:", length(channels)))
```

### Select channels to use for clustering

Reduced dimension plots showing marker expression that generated by the previous script 
can be used to select the most relevant markers for clustering.

In the first step, we just use Cell Category markers for clustering.

```{r select-channels-clustering}
channels_clust <- rownames(rowData(spe)[rowData(spe)$clustering == 1, ])
cat(c("\nChannels used for unsupervised clustering:",
      channels_clust[channels_clust %in% rownames(spe)]))

# Select Channels we will do our Category assignment on.
channels_cat <- c("SYP", "AMY", "CK19", 
                  "CD45RA", "CD45RO", "MPO", "CD68", 
                  "SMA", "CD31")
cat(c("\nChannels used for cell category assignment:",
      channels_cat[channels_cat %in% rownames(spe)]))
```


## Subset

```{r convert-to-sce}
# Extract the metadata (to avoid conflicts when merging the SCEs)
meta <- metadata(spe)
spatial_coords <- spatialCoords(spe)
colpairs <- colPairs(spe)

# Convert to SingleCellExperiment
sce <- as(spe, "SingleCellExperiment")
# sce_cat <- sce[channels_cat, ]
sce_cat <- spe[channels_cat, ]
```

## **Clustering: PhenoGraph**

Unsupervised clustering is performed with the [PhenoGraph](https://doi.org/10.1016/j.cell.2015.05.047) algorithm. 
This method works by generating a nearest-neighbor (kNN) graph of phenotpyic similarities followed by Louvain community dectection. 

Here, we try the [Rphenoannoy implementation](https://github.com/stuchly/Rphenoannoy).
RPhenoannoy implements a parallel Jaccard-coefficient, approximates the kNN and uses Louvain clustering.

```{r phenograph-run}
clust_method <- c("Pheno1")
cur_assay <- "scaled"
# Number of nearest-neighbors
k <- 30
# Run Phenograph - PhenoGraph for exprs and scaled assay.
clust_name <- paste(clust_method, cur_assay, "cat", sep = "_")
writeLines(c("\n", clust_name))

if (!clust_name %in% colnames(colData(sce_cat))) {
  set.seed(seed)
  # Run Rphenograph.
  cur_pheno_annoy <- Rphenoannoy::Rphenoannoy(t(assay(sce_cat, cur_assay)), k = k)

  cur_pheno <- DataFrame(cur_pheno_annoy[[2]]$membership)
  colnames(cur_pheno) <- clust_name
  rownames(cur_pheno) <- colnames(assay(sce_cat, cur_assay))

  # Add Phenograph clusters to the colData of the SCE object
  # Cluster `0` is attributed to non-subsetted cells and not islet cells
  colData(sce_cat)[, clust_name] <- cur_pheno
  remove(cur_pheno)
}
```

# **Visualize clusters**

## Load packages

```{r packages3, results="hide"}
suppressPackageStartupMessages(c(
  library(heatmaply),
  library(htmltools)#,
  #library(cytomapper)
))
```

Subset SCE-CAT object.
```{r}
set.seed(222)
if (!("subset" %in% names(metadata(sce_cat)))) {
  # Cells per case
  nb_cells <- 7500
  
  # Subset the SPE object (nb_cells per case)
  cell_subset <- tibble(rn = rownames(colData(sce_cat)),
                        case_id = colData(spe)$case_id) |> 
    group_by(case_id) |>
    sample_n(nb_cells) |> 
    pull(rn)
  
  # Keep the subset cell ids in SPE metadata
  metadata(spe)[["subset"]] <- sort(as.vector(cell_subset))
}

# Subset the SPE object
sce_sub_cat <- sce_cat[channels_cat, metadata(spe)[["subset"]]]
```

## Visualize clusters on reduced dimensions
RUN UMAP on reduced category SPE.
```{r dimred-clusters-categories}
# Save as variable and write assays to NULL. Reduces memory overhead for parallelization. 
# Add UMAPs to SPE object
dimred_name <- paste("UMAP", cur_assay, "cat", sep = "_")
print(dimred_name)

# Run UMAP on a cell subset
if ((!dimred_name %in% reducedDimNames(sce_sub_cat)) && ("UMAP" %in% dimred_sel)) {
  # Extract Counts. 
  if (cur_assay %in% assayNames(spe)) {
    counts <- t(assay(sce_sub_cat, cur_assay))
  }
  # Run UMAP.
  umap_model <- uwot::umap(counts, ret_model = TRUE)
  # Extract Embedding.
  cur_umap <- umap_model$embedding
  colnames(cur_umap) <- c("UMAP1", "UMAP2")
  rownames(cur_umap) <- rownames(counts)
}
reducedDim(sce_sub_cat, dimred_name) <- cur_umap
```

```{r dimred-clusters}
cur_method <- "Pheno1"
cur_dimred <- "UMAP"
cur_assay <- "scaled"

cur_dat <- makePerCellDF(sce_sub_cat, use_dimred = TRUE) |>
  dplyr::mutate(case_id = factor(case_id, levels = meta$cases),
                donor_type = factor(donor_type, levels = meta$stages)) |>
  dplyr::arrange(case_id, donor_type)  |> 
  tibble::as_tibble()

dimred_name <- paste(cur_dimred, cur_assay, "cat", sep = "_")
clust_name <- paste(cur_method, cur_assay, "cat", sep = "_")

# Plot all clustersdat, dimred, color_by
p <- plot_dim_red(dat = cur_dat, dimred = dimred_name, color_by = clust_name,
                  sample = TRUE, size = 0.1, alpha = 1)
print(p)
fn <- paste0(paste(today, "Clusters", clust_name, cur_dimred,
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

## Heatmap of marker expression by cluster
```{r heatmap-clusters-markers}
name_cur_assay <- "scaled"
clust_name <- paste(cur_method, cur_assay, "cat", sep = "_")
message(clust_name)
message(name_cur_assay)
# Summarize the data

hm <- summarize_heatmap(sce_cat,
                        expr_values = name_cur_assay,
                        cluster_by = clust_name,
                        channels = channels_cat)

# Display the heatmap
fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap",
                    sep = "_"), ".html")
print(fn)
heatmaply::heatmaply(
    heatmaply::normalize(hm), main = clust_name, 
    file = file.path(paths$folder_script, fn))

# remove all clusters with less than 10 cells and plot again.
clust_freq <- table(colData(sce_cat)[[clust_name]])
keep_clusts <- names(clust_freq)[clust_freq >= 10]
sce_isl_sub <- sce_cat[, colData(sce_cat)[[clust_name]] %in% keep_clusts]

hm <- summarize_heatmap(sce_isl_sub,
                        expr_values = name_cur_assay,
                        cluster_by = clust_name,
                        channels = channels_cat)
# Display the reduced heatmap -> this improves the visualization.
# Otherwise heatmap is scaled by expression of outliers.
fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap_reduced",
                    sep = "_"), ".html")
print(fn)
heatmaply::heatmaply(
    heatmaply::normalize(hm), main = clust_name, 
    file = file.path(paths$folder_script, fn))
```


Put these into the following compartments:
Islet, Immune, Exocrine, Stroma, Other.

## Visualize clusters with cytomapper and cytoviewer

Here, we use the `cytoviewer` package to visualize the clusters on the images.
Thereby, we can check if the clusters are biologically meaningful and if they indeed correspond to the respective cell types.

Additional information we can use is their spatial localization of cell types, i.e. which cells in the endocrine or exocrine compartment.

### Select cluster(s) and assay to show

```{r viz-clust-select}
viz_clust <- c(9, 15, 8, 3, 6, 5, 2) # Select cluster(s) to visualize.
viz_method <- "Pheno1"
viz_assay <- "scaled"
```


Adjust function.
```{r}
imgloader <- function(x, image_dir, image_names,
                     suffix_rem = "", suffix_add = "",
                     bit_depth = 16, type, ...) {
  require(cytomapper)

  image_list <- file.path(image_dir, image_names)

  # Test if the image list exist
  test_exist <- which(!file.exists(image_list))
  if (length(test_exist) > 0) {
    stop(c("The following images were not found:\n",
           paste(image_list[test_exist], collapse = "\n")))
  } else {
    # Load and scale the images
    images <- loadImages(image_list, ...)
    # images <- scaleImages(images, (2 ^ bit.depth) - 1)

    # Add image names to metadata
    mcols(images)$ImageName <- gsub(suffix_rem, "", names(images))
    mcols(images)$ImageName <- paste0(mcols(images)$ImageName, suffix_add)

    # Add channel names
    if (type == "stacks") {
      print("Loading image stacks")
      channelNames(images) <- rownames(x)
    }
    return(images)
  }
}
```


### Load images and masks

```{r viz-cluster-load}
spe$Pheno1_scaled_cat <- sce_cat$Pheno1_scaled_cat

if (!is.null(viz_clust)) {
  nb_images <- 14
  image_extension <- ".tiff"
  clust_name <- paste(viz_method, viz_assay, "cat", sep = "_")

  # Subset the SCE
  sce_viz <- spe[, colData(spe)[[clust_name]] %in% viz_clust]

  # Select random image.
  set.seed(seed)
  image_sub <- sort(sample(
    unique(sce_viz$image_fullname),
  min(length(unique(sce_viz$image_fullname)), nb_images)))

  # Folders
  folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
  folder_masks <- file.path(paths$folder_in, "masks_cells",
                            paths$panel_type, "whole-cell")

  # Load images and masks
  images <- imgloader(
    x = sce_viz,
    image_dir = folder_images,
    image_names = image_sub,
    type = "stacks"
  )

  masks <- imgloader(
    x = sce_viz,
    image_dir = folder_masks,
    image_names = image_sub,
    as.is = TRUE,
    type = "masks"
  )

  sce_viz <- sce_viz[, sce_viz$image_fullname %in% image_sub]
  sce_viz$ImageName <- gsub(image_extension, "", sce_viz$image_fullname)
}
```

### Interactive with Cytoviewer

```{r viz-cluster-cytoviewer}
# Use cytoviewer with images, masks and object
library(cytoviewer)
if (!is.null(viz_clust)) {
  channels_view <- channels_clust
  sub_images <- cytomapper::getChannels(images, channels_view)
  app <- cytoviewer(image = sub_images, 
                    mask = masks, 
                    object = sce_viz[channels_view, ], 
                    img_id = "ImageName", 
                    cell_id = "cell_number")

  if (interactive()) {
    shiny::runApp(app)
  }
}
```



# **Attribute cell categories**

**Note: this section requires manual intervention**

Here, clusters obtained at the previous steps are manually merged into meaningful cell categories.
Cluster numbers are attributed to the different cell categories based on the plots and heatmaps above. 
Consequently, this attribution has to be adapted to the clustering results.

## Cell category attribution

### Phenograph x scaled counts

```{r cellcats-pheno-scaled}
clust_methods <- c("Pheno1")
clust_assay <- c("scaled")

if (!clust_assay %in% assay_sel) stop("The selected assay is not in the assays selected for clustering")
if (length(clust_assay) != 1) stop("Select only one assay")

clust_name <- paste(clust_methods, clust_assay, "cat", sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
# 
# Mesenchymal: 28, 1, 4, 26,
# Exocrine: 26, 27, 25, 23, 7, 14, 17, 11, 13, 6, 5, 2, 3.
# Islet: 21, 18, 19, 10.
# Immune: 29, 22, 24, 12.
# Other: 9, 15, 8.

clust_other <- c(9, 15, 8)
clust_islet <- c(21, 18, 19, 10)
clust_immune <- c(29, 22, 24, 12)
clust_stroma <- c(28, 1, 4, 16, 20)
clust_exocrine <- c(26, 27, 25, 23, 7, 14, 17, 11, 13, 6, 5, 2, 3)

all_clust <- sort(c(clust_islet, clust_immune, clust_exocrine, clust_stroma,clust_other))  

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce_cat)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
# Islet
colData(sce_cat)[colData(sce_cat)[, clust_name] %in% clust_islet,
                "CellCat"] <- "Islet"
colData(sce_cat)[colData(sce_cat)[, clust_name] %in% clust_immune,
                "CellCat"] <- "Immune"
colData(sce_cat)[colData(sce_cat)[, clust_name] %in% clust_exocrine,
                "CellCat"] <- "Exocrine"
colData(sce_cat)[colData(sce_cat)[, clust_name] %in% clust_stroma,
                "CellCat"] <- "Stroma"
colData(sce_cat)[colData(sce_cat)[, clust_name] %in% clust_other,
                "CellCat"]<- "Other"
```

### Plot cell types on reduced dimensions

```{r cellcats-dimred}
sce_sub_cat$Pheno1_scaled_cat <- sce_cat[, metadata(spe)[["subset"]]]$Pheno1_scaled_cat
sce_sub_cat$CellCat <- sce_cat[, metadata(spe)[["subset"]]]$CellCat

# Prepare the data
cur_dat <- makePerCellDF(sce_sub_cat, use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = metadata(sce_sub_cat)$cases),
         donor_type = factor(donor_type, levels = metadata(sce_sub_cat)$stages)) |>
  arrange(case_id, donor_type)

# Plot
cur_assay <- "scaled"
cur_dimred <- "UMAP"
cur_method <- "Pheno1"

dimred_name <- paste(cur_dimred, cur_assay, "cat", sep = "_")
clust_name <- "CellCat"

p <- plot_dim_red(cur_dat, dimred_name, clust_name,
                  sample = TRUE, size = 1, alpha = 1)
print(p)

fn <- paste0(paste(today, clust_name, cur_dimred,
                  sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

### Cell categories heatmap

```{r cellcats-heatmap}
name_cur_assay <- "scaled"

# Summarize the data
hm <- summarize_heatmap(sce_cat,
                        expr_values = name_cur_assay,
                        cluster_by = clust_name,
                        channels = channels_cat)

# Display the heatmap
fn <- paste0(paste(today, clust_name, "Heatmap",
                sep = "_"), ".html")

heatmaply(
heatmaply::normalize(hm), main = clust_name,
file = file.path(paths$folder_script, fn))
```


### Number of cells by cell category

```{r cells-per-category}
table(colData(sce_cat)[, "CellCat"])
```

# **Save **

```{r spe-transfer-categories}
sce_cat <- sce_cat[, colnames(spe)]
spe$cell_category <- colData(sce_cat)$CellCat
table(spe$cell_category)
```

```{r spe-transfer-categories2}
fn_spe <- file.path(paths$folder_out, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
print(spe)
saveRDS(spe, fn_spe)
```