---
title: "05_CellCategories_cells_Compressed"
author: "Nathan Steenbuck"
date: "Created: 23 Oct, 2024; Compiled: `r format(Sys.time(), '%d %b, %')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
script_name <- "05_CellCategories_cells_Uncompressed.Rmd"

source(file.path("/", "mnt", "central_nas", "projects", 
                 "type1_diabetes", "nathan", "BlockCourse", "2024_BlockCourse", "T1D_analysis", 
                "helpers.R"))
n_cores <- 2
future::plan(future::multicore(workers = n_cores))
paths <- getPaths(script_name)
knitr::opts_knit$set(root_dir = paths$cluster_home)
do_print <- TRUE

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("04_QualityControl", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```



# **Goal**

In the following scripts, cell types are attributed to all cells in the dataset
in an iterative way:

- `05_CellCategories_cells_Uncompressed` (this script): Attribution of cell types from the compressed dataset.

This is performed by performing PhenoGraph clustering using the Rphenoannoy package.

For cell type annotation we perform unsupervised clustering, and then relate the lineage marker expression of clusters to cell types.

The resulting cell categories, which are used in downstream analyses are stored
as `colData(spe)$cell_category`.

# **Settings**

## Load packages

```{r packages, results="hide"}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(parallel),
  library(tictoc),
  library(purrr),
  library(furrr)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```



# **Clustering**

## Settings

### Load packages

RPhenoannoy can installed with: `devtools::install_github("stuchly/Rphenoannoy@8b81e2e7fb0599f45070e2cba1b28ac219b7c472")`

```{r packages2, results="hide"}
suppressPackageStartupMessages(c(
  library(ggplot2),
  library(scater),
  library(scuttle),
  library(scran),
  library(igraph),
  library(Rphenoannoy),
  library(clustree),
  library(ranger),
  library(patchwork),
  library(BiocParallel),
  library(tictoc)
))
```

### Channels and assays

Select channels (`channels_clust`), assays (`assay_sel`) and clustering methods (`methods_sel`) to use.

```{r select-methods-assays}
methods_sel <- c("Pheno1")

assay_sel <- c("scaled")
names(assay_sel) <- c("scaled")

writeLines(c("Assays:", assay_sel[assay_sel %in% assayNames(spe)], 
                        assay_sel[assay_sel %in% reducedDimNames(spe)]))

dimred_sel <- c("UMAP")
writeLines(c("\nReduced dimensions:", dimred_sel))

channels <- rownames(spe)[!(grepl("DNA|H3", rownames(spe)))]
cat(c("\nChannels:", channels[channels %in% rownames(spe)]))
cat(c("\nNumber of channels:", length(channels)))
```

### Select channels to use for clustering

Reduced dimension plots showing marker expression that generated by the previous script 
can be used to select the most relevant markers for clustering.

```{r select-channels-clustering}
channels_clust <- rownames(rowData(spe)[rowData(spe)$clustering == 1, ])
cat(c("\nChannels used for unsupervised clustering:",
      channels_clust[channels_clust %in% rownames(spe)]))
```


## Subset

```{r convert-to-sce}
# Extract the metadata (to avoid conflicts when merging the SCEs)
meta <- metadata(spe)
spatial_coords <- spatialCoords(spe)
colpairs <- colPairs(spe)

# Convert to SingleCellExperiment
sce <- as(spe, "SingleCellExperiment")
remove(spe)
```

## **Clustering: PhenoGraph**

Unsupervised clustering is performed with the [PhenoGraph](https://doi.org/10.1016/j.cell.2015.05.047) algorithm. 
This method works by generating a nearest-neighbor (kNN) graph of phenotpyic similarities followed by Louvain community dectection. 

Here, we try the [Rphenoannoy implementation](https://github.com/stuchly/Rphenoannoy).
RPhenoannoy implements a parallel Jaccard-coefficient, approximates the kNN and uses Louvain clustering.

```{r phenograph-run}
clust_method <- c("Pheno1")

# Number of nearest-neighbors
k <- 30

# Run Phenograph - PhenoGraph for exprs and scaled assay.
purrr::map(assay_sel, \(cur_assay) {
  clust_name <- paste(clust_method, cur_assay, sep = "_")
  writeLines(c("\n", clust_name))

  if (!clust_name %in% colnames(colData(sce))) {
    set.seed(seed)
    # Run Rphenograph.
    cur_pheno_annoy <- Rphenoannoy::Rphenoannoy(t(assay(sce, cur_assay)), k = k)

    cur_pheno <- DataFrame(cur_pheno_annoy[[2]]$membership)
    colnames(cur_pheno) <- clust_name
    rownames(cur_pheno) <- colnames(assay(sce, cur_assay))

    # Add Phenograph clusters to the colData of the SCE object
    # Cluster `0` is attributed to non-subsetted cells and not islet cells
    colData(sce)[, clust_name] <<- cur_pheno
    remove(cur_pheno)
  }
})
```

# **Visualize clusters**

## Load packages

```{r packages3, results="hide"}
suppressPackageStartupMessages(c(
  library(heatmaply),
  library(htmltools)#,
  #library(cytomapper)
))
```


## Visualize clusters on reduced dimensions

```{r dimred-clusters}
cur_method <- "Pheno1"
cur_dimred <- "UMAP"
cur_assay <- "scaled"
cur_dat <- makePerCellDF(sce, use_dimred = TRUE) |>
  dplyr::mutate(case_id = factor(case_id, levels = meta$cases),
                donor_type = factor(donor_type, levels = meta$stages)) |>
  dplyr::arrange(case_id, donor_type)  |> 
  tibble::as_tibble()

dimred_name <- paste(cur_dimred, cur_assay, sep = "_")
clust_name <- paste(cur_method, cur_assay, sep = "_")

# Plot all clustersdat, dimred, color_by
p <- plot_dim_red(dat = cur_dat, dimred = dimred_name, color_by = clust_name,
                  sample = TRUE, size = 0.1, alpha = 1)
if (do_print) print(p)
fn <- paste0(paste(today, "Clusters", clust_name, cur_dimred,
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```


## Distribution of marker expression by cluster

```{r violin-clusters-markers, message=FALSE, warning=FALSE, fig.width=15, fig.height=10}
clust_name <- paste(cur_method, cur_assay, sep = "_")
print(clust_name)
sce[[clust_name]] <- as.factor(sce[[clust_name]])

if (cur_assay %in% SummarizedExperiment::assayNames(sce)) {
    cur_dat <- scuttle::makePerCellDF(sce, assay.type = cur_assay,
                                      features = channels_clust, use.dimred = FALSE)
}

cur_dat <- melt.data.table(
  as.data.table(cur_dat),
  id.vars = c("image_number", "cell_number", clust_name),
  measure.vars = channels_clust,
  variable.name = "channel",
  value.name = cur_assay
)
cur_dat[, MeanExprs := mean(get(cur_assay)),
        by = c("channel", clust_name)]

p <- plot_violin(cur_dat, x = "channel", y = cur_assay,
                  fill_by = "MeanExprs", facet_by = clust_name,
                  title = clust_name) +
  theme(axis.text.x = element_text(size = 8, angle = 90))
if (do_print) suppressMessages(print(p))
fn <- paste0(paste(today, "Clusters", clust_name, "Violins",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```


## Heatmap of marker expression by cluster
```{r heatmap-clusters-markers}
hmlist <- list()
n <- 1
name_cur_assay <- "scaled"
clust_name <- paste(cur_method, cur_assay, sep = "_")
message(clust_name)
message(name_cur_assay)
# Summarize the data

hm <- summarize_heatmap(sce,
                        expr_values = name_cur_assay,
                        cluster_by = clust_name,
                        channels = channels_clust)

# Display the heatmap
fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap",
                    sep = "_"), ".html")
print(fn)
hmlist[[n]] <- heatmaply(
    heatmaply::normalize(hm), main = clust_name, 
    file = file.path(paths$folder_script, fn))
n <- n + 1

# remove all clusters with less than 10 cells and plot again.
clust_freq <- table(colData(sce)[[clust_name]])
keep_clusts <- names(clust_freq)[clust_freq >= 10]
sce_isl_sub <- sce[, colData(sce)[[clust_name]] %in% keep_clusts]

hm <- summarize_heatmap(sce_isl_sub,
                        expr_values = name_cur_assay,
                        cluster_by = clust_name,
                        channels = channels_clust)
# Display the reduced heatmap -> this improves the visualization.
# Otherwise heatmap is scaled by expression of outliers.
fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap_reduced",
                    sep = "_"), ".html")
print(fn)
hmlist[[n]] <- heatmaply(
    heatmaply::normalize(hm), main = clust_name, 
    file = file.path(paths$folder_script, fn))
n <- n + 1

htmltools::tagList(setNames(hmlist, NULL))
```


## Visualize clusters with cytomapper and cytoviewer

Here, we use the `cytoviewer` package to visualize the clusters on the images.
Thereby, we can check if the clusters are biologically meaningful and if they indeed correspond to the respective cell types.

Additional information we can use is their spatial localization of cell types, i.e. which cells in the endocrine or exocrine compartment.

### Select cluster(s) and assay to show

```{r viz-clust-select}
viz_clust <- NULL # Select cluster(s) to visualize.
viz_method <- "Pheno1"
viz_assay <- "scaled"
```

### Load images and masks

```{r viz-cluster-load}
if (!is.null(viz_clust)) {
  nb_images <- 14
  image_extension <- ".tiff"
  clust_name <- paste(viz_method, viz_assay, sep = "_")

  # Subset the SCE
  sce_viz <- sce[, colData(sce)[[clust_name]] %in% viz_clust]

  # Select random image.
  set.seed(seed)
  image_sub <- sort(sample(
    unique(sce_viz$image_fullname),
  min(length(unique(sce_viz$image_fullname)), nb_images)))

  # Folders
  folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
  folder_masks <- file.path(paths$folder_in, "masks_cells",
                            paths$panel_type, "whole-cell")

  # Load images and masks
  images <- yoloader(
    x = sce_viz,
    image_dir = folder_images,
    image_names = image_sub,
    type = "stacks"
  )

  masks <- yoloader(
    x = sce_viz,
    image_dir = folder_masks,
    image_names = image_sub,
    as.is = TRUE,
    type = "masks"
  )

  sce_viz <- sce_viz[, sce_viz$image_fullname %in% image_sub]
  sce_viz$ImageName <- gsub(image_extension, "", sce_viz$image_fullname)
}
```

### Interactive with Cytoviewer

```{r viz-cluster-cytoviewer}
# Use cytoviewer with images, masks and object
# library(cytoviewer)
if (!is.null(viz_clust)) {
  channels_view <- channels_clust
  sub_images <- cytomapper::getChannels(images, channels_view)
  app <- cytoviewer(image = sub_images, 
                    mask = masks, 
                    object = sce_viz[channels_view, ], 
                    img_id = "ImageName", 
                    cell_id = "cell_number")

  if (interactive()) {
    shiny::runApp(app)
  }
}
```



# **Attribute cell categories**

**Note: this section requires manual intervention**

Here, clusters obtained at the previous steps are manually merged into meaningful cell types.
Cluster numbers are attributed to the different cell types based on the plots and heatmaps above. 
Consequently, this attribution has to be adapted to the clustering results.

## Cell category attribution

Cell category attribution has to be performed for every combination of clustering methods and selected assays.

### Phenograph x scaled counts

Depending on the clustering, maybe select other categories.

```{r cellcats-pheno-scaled}
clust_methods <- c("Pheno1")
clust_assay <- c("scaled")

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")
if (length(clust_assay) != 1)
  stop("Select only one assay")

clust_name <- paste(clust_methods, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
clust_alpha <- c(17,11)
clust_beta  <- c(16)
clust_delta <- c(13)
clust_gamma <- c(22)

# Immune cells.
clust_lympho <- c(7)
clust_myeloid <- c(6) # CD68
clust_neutrophil <- c() # MPO?

# Exocrine Cells.
clust_ductal <- c(3, 12, 20)
clust_acinar <- c(19, 8, 9, 2, 14, 5, 15)
clust_endothelial <- c(1) #CD31
clust_mesenchymal <- c()
clust_other <- c(10,18,4,21) #

all_clust <- sort(c(clust_alpha, clust_beta, clust_delta, clust_gamma,
                    clust_lympho, clust_myeloid, clust_neutrophil,
                    clust_ductal, clust_acinar, clust_endothelial, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
# Islet
colData(sce)[colData(sce)[, clust_name] %in% clust_alpha,
             "CellType"] <- "Alpha"
colData(sce)[colData(sce)[, clust_name] %in% clust_beta,
              "CellType"] <- "Beta"
colData(sce)[colData(sce)[, clust_name] %in% clust_delta,
              "CellType"] <- "Delta"
colData(sce)[colData(sce)[, clust_name] %in% clust_gamma,
               "CellType"]<- "Gamma"
# Immune:
colData(sce)[colData(sce)[, clust_name] %in% clust_lympho,
             "CellType"] <- "Lympho"
colData(sce)[colData(sce)[, clust_name] %in% clust_myeloid,
             "CellType"] <- "Myeloid"
colData(sce)[colData(sce)[, clust_name] %in% clust_neutrophil,
              "CellType"] <- "Neutrophil"
# Exocrine
colData(sce)[colData(sce)[, clust_name] %in% clust_ductal,
              "CellType"] <- "Ductal"
colData(sce)[colData(sce)[, clust_name] %in% clust_acinar,
              "CellType"] <- "Acinar"
# Stroma:
colData(sce)[colData(sce)[, clust_name] %in% clust_endothelial,
              "CellType"] <- "Endothelial"
colData(sce)[colData(sce)[, clust_name] %in% clust_mesenchymal,
              "CellType"] <- "Fibro_Endo"
# Other
colData(sce)[colData(sce)[, clust_name] %in% clust_other,
              "CellType"] <- "Other"
```

Assign Cell Categories base don the assigned cell types.
```{r cell_cats}
# Islet-category (Alpha, Beta, Delta, Gamma)
colData(sce)[colData(sce)[, "CellType"] %in% c("Alpha", "Beta", "Delta", "Gamma"),
             "CellCat"] <- "Islet"
# Immune-category (Lympho, Myeloid, Neutrophil)
colData(sce)[colData(sce)[, "CellType"] %in% c("Lympho", "Myeloid", "Neutrophil"),
             "CellCat"] <- "Immune"
# Exocrine-category (Ductal, Acinar)
colData(sce)[colData(sce)[, "CellType"] %in% c("Ductal", "Acinar"),
              "CellCat"] <- "Exocrine"
# Stroma-category (Endothelial, Fibro_Endo)
colData(sce)[colData(sce)[, "CellType"] %in% c("Endothelial", "Mesenchymal"),
              "CellCat"] <- "Stroma"
# Other-category (Other)
colData(sce)[colData(sce)[, "CellType"] %in% c("Other"),
              "CellCat"] <- "Other"
```

### Plot cell types on reduced dimensions

```{r cellcats-dimred}
# Prepare the data
cur_dat <- makePerCellDF(sce, use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = metadata(sce)$cases),
         donor_type = factor(donor_type, levels = metadata(sce)$stages)) |>
  arrange(case_id, donor_type)

# Plot
cur_assay <- "scaled"
cur_dimred <- "UMAP"
cur_method <- "Pheno1"

dimred_name <- paste(cur_dimred, cur_assay, sep = "_")
clust_name <- "CellType"

p <- plot_dim_red(cur_dat, dimred_name, clust_name,
                  sample = TRUE, size = 1, alpha = 1)
if (do_print) { print(p)}

fn <- paste0(paste(today, clust_name, cur_dimred,
                  sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

### Cell categories heatmap

```{r cellcats-heatmap}
cur_assay <- "scaled"
name_cur_assay <- "scaled"
clust_name <- "CellType"

# Summarize the data
hm <- summarize_heatmap(sce,
                    expr_values = name_cur_assay,
                    cluster_by = clust_name,
                    channels = channels_clust)

# Display the heatmap
fn <- paste0(paste(today, clust_name, "Heatmap",
                sep = "_"), ".html")

heatmaply(
heatmaply::normalize(hm), main = clust_name,
file = file.path(paths$folder_script, fn))
```


### Number of cells by cell category

```{r cells-per-category}
table(colData(sce)[, "CellCat"])
table(colData(sce)[, "CellType"])
```


# **Plot cell categories**

## Reduced dimensions

Plot consensus cell categories on reduced dimensions

```{r consensus-dimred}
ccat <- c("Immune", "Islet", "Exocrine", "Stroma", "Other")
names(ccat) <- c("Immune", "Islet", "Exocrine", "Stroma", "Other")
palette_ccat <- c(palettes$colors[1:(length(ccat) - 2)], "grey40", "grey")
names(palette_ccat) <- ccat

cur_dat <- scuttle::makePerCellDF(sce[, colnames(sce) %in% metadata(sce)$subset],
                                  use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = metadata(sce)$cases),
         donor_type = factor(donor_type, levels = metadata(sce)$stages)) |>
  arrange(case_id, donor_type)


dimred_name <- paste(cur_dimred, cur_assay, sep = "_")

p <- plot_dim_red(cur_dat, dimred_name, "CellCat",
                  sample = TRUE, size = 0.1, alpha = 1,
                  palette = palette_ccat)
if (do_print) print(p)

fn <- paste0(paste(today, "CellCat", "Consensus", dimred_name,
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

## Heatmap

```{r consensus-heatmap}
hmlist <- list()

purrr::map(assay_sel, \(cur_assay) {
  clust_name <- paste("cell_category", cur_assay, sep = "_")

  # Summarize the data
  hm <- summarize_heatmap(sce,
                          expr_values = cur_assay,
                          cluster_by = "cell_category",
                          channels = channels)

  # Display the heatmap
  fn <- paste0(paste(today, "CellCat", "Consensus", cur_assay,
                     "Heatmap", sep = "_"), ".html")

  hmlist[[i]] <- heatmaply(
    heatmaply::normalize(hm), main = clust_name,
    file = file.path(paths$folder_script, fn))
})
htmltools::tagList(setNames(hmlist, NULL))
```


## Correlation between channels

```{r cell-category-correlation}
# Define the correlation method
cor_method <- "pearson"

# Function to calculate p values
cor_test_p <- function(x) {
  FUN <- function(x, y) cor.test(x, y, method = cor_method)[["p.value"]]
  z <- outer(
    colnames(x),
    colnames(x),
    Vectorize(function(i, j) FUN(x[, i], x[, j]))
  )
  dimnames(z) <- list(colnames(x), colnames(x))
  z
}

hmlist <- list()
purrr::map2(assay_sel, names(assay_sel), \(cur_assay, name_cur_assay) {

  # Summarize the data
  hm <- summarize_heatmap(sce,
                          expr_values = name_cur_assay,
                          cluster_by = "cell_category",
                          channels = channels)

  # Run the correlation test
  r <- cor(hm, method = cor_method)
  p <- cor_test_p(hm)
  diag(p) <- NA

  # Plot and save the heatmap
  fn <- paste0(paste(today, "CellCat", "Consensus", "Correlation",
                     cur_assay, sep = "_"), ".html")

  hmlist[[i]] <- heatmaply::heatmaply_cor(
    r, node_type = "scatter",
    point_size_mat = -log10(p), point_size_name = "-log10(p-value)",
    label_names = c("x", "y", "Correlation"),
    main = paste("Cell type -", cur_assay),
    file = file.path(paths$folder_script, fn))
})
htmltools::tagList(setNames(hmlist, NULL))
```

# **Save **

## Remove temporary columns

```{r remove-cellcat-columns}
cellcat_cols <- colnames(colData(sce))[grep("CellCat_", colnames(colData(sce)))]
colData(sce)[, cellcat_cols] <- NULL
```

## Convert to SPE
