---
title: "07_CellTypesComparison_cells_Compressed"
author: "Nathan Steenbuck"
date: "Created: 23 Oct, 2024; Compiled: `r format(Sys.time(), '%d %b, %')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

Rscript -e "rmarkdown::render('2024_BlockCourse/T1D_analysis/07_CellTypesComparison_cells_Compressed.Rmd')"

# **Goal**

- Cell Type attributions between the Compressed and Uncompressed datasets are compared in this script.

## Steps:
- Read in both datasets.
- Compare the cell type attributions between the two datasets per ROI.
- Visualize the differences.
- Calculate correlations
- If time: compare potential beta-cell loss between stages, or increase in T-Cells (CD4+/CD8+)


```{r setup, include=FALSE}
script_name <- "07_CellTypesComparison_cells_Compressed.Rmd"

source(file.path("/", "mnt", "central_nas", "projects", 
                 "type1_diabetes", "nathan", "BlockCourse", "2024_BlockCourse", "T1D_analysis", 
                "helpers.R"))
n_cores <- 2
future::plan(future::multicore(workers = n_cores))
paths <- getPaths(script_name)
knitr::opts_knit$set(root_dir = paths$cluster_home)
do_print <- TRUE

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("06_CellTypes", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```

# **Settings**

## Load packages

```{r packages, results="hide"}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(parallel),
  library(tictoc),
  library(purrr),
  library(furrr)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Load the SpatialExperiment (SPE) object from the compressed dataset.

```{r load-data-compressed}
fn_spe_compressed <- file.path(paths$folder_out, paste0(paths$object_type, "_", "Compressed.rds"))
spe_compressed <- readRDS(fn_spe_compressed)
print(spe_compressed)

compressed_df <- scuttle::makePerCellDF(spe_compressed, use.dimred = FALSE) |> 
  mutate(method = "Compressed") |> 
  select(case_id, panel, cell_number, cell_type, cell_category, method, image_id)
```

Load the SpatialExperiment (SPE) object from the uncompressed dataset.

```{r load-data2}
fn_spe_uncompressed <- file.path(paths$folder_out, paste0(paths$object_type, "_", "Uncompressedsce.rds"))
spe_uncompressed <- readRDS(fn_spe_uncompressed)
print(spe_uncompressed)

uncompressed_df <- scuttle::makePerCellDF(spe_uncompressed, use.dimred = FALSE) |> 
  mutate(method = "Uncompressed") |> 
  select(case_id, panel, cell_number, cell_type, cell_category, method, image_id)
```

Comparison of total cell numbers per Cell Types between the two datasets.

```{r barplot-all-cts}
# Dodged Barplot of Cell Types
df <- compressed_df  |> 
  bind_rows(uncompressed_df) |> 
  mutate(method = factor(method, levels = c("Compressed", "Uncompressed"))) |> 
  as_tibble()

p <- df |> 
  dplyr::count(cell_type, method) |>
  ggplot(aes(x = cell_type, y = n, fill = method)) +
  geom_bar(stat = "identity", position = "dodge") +
  mytheme$standard()

print(p)
fn <- file.path(paths$folder_script, paste0(today, "_CellTypes_Comparison_Compressed_Uncompressed", ".png"))
do.call(ggsave, c(list(fn, p), plotsave_param))
```

Reduced plot for the most important cell types.
```{r barplot-beta-t}
p <- df |> 
  filter(cell_type %in% c("Beta", "T_CD4", "T_CD8", "T_cell")) |> 
  mutate(cell_type = ifelse(cell_type %in% c("T_CD4", "T_CD8"), "T_cell", cell_type)) |>
  dplyr::count(cell_type, method, case_id) |>
  ggplot(aes(x = case_id, y = n, fill = method)) +
  geom_bar(stat = "identity", position = "dodge") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~cell_type, scales = "free_y") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)

fn <- file.path(paths$folder_script, paste0(today, "_CellTypes_Comparison_Compressed_Uncompressed_Reduced", ".png"))
do.call(ggsave, c(list(fn, p), plotsave_param))
```

Now, we will calculate the correlation between the cell types per ROI.

```{r correlations-t-beta-per-roi}
# Visualize Cell-Category Correlation per ROI.
cor_df <- df |> 
  mutate(image_id = stringr::str_remove(image_id, "_Compressed")) |> 
  mutate(image_id = stringr::str_remove(image_id, "_Uncompressed")) |>
  mutate(cell_type = ifelse(cell_type %in% c("T_CD4", "T_CD8"), "T_cell", cell_type)) |>
  dplyr::count(method, case_id, cell_type, image_id) |> # Count occurrences
  tidyr::pivot_wider(names_from = method, values_from = n, values_fill = 0)

p <- cor_df |> 
  filter(cell_type %in% c("Beta")) |> 
  ggplot(aes(x = Compressed, y = Uncompressed)) +
  geom_point() +
  geom_smooth(method = "lm") +
  mytheme$standard_new() +
  labs(title = "",
       x = "Compressed: Beta-cell Number", y = "Uncompressed: Beta-cell Number")
print(p)

fn <- file.path(paths$folder_script, paste0(today, "_Correlation_Compressed_Uncompressed_Beta",  ".png"))
do.call(ggsave, c(list(fn, p), plotsave_param))

p <-  cor_df |> 
  filter(cell_type %in% c("T_cell")) |> 
  ggplot(aes(x = Compressed, y = Uncompressed)) +
  geom_point() +
  geom_smooth(method = "lm") +
  mytheme$standard_new() +
  labs(title = "",
       x = "Compressed: T-cell Number", y = "Uncompressed: T-cell Number")
print(p)

fn <- file.path(paths$folder_script, paste0(today, "_Correlation_Compressed_Uncompressed_Beta",  ".png"))
do.call(ggsave, c(list(fn, p), plotsave_param))
```

Show correlations for all Cell types.
```{r correlation2}
# Calculate Cell-Type Correlation per ROI.
p <- cor_df |> 
  #filter(cell_type != "Fibro_SM") |> 
  filter(cell_type != "Other") |>
  group_by(cell_type) |> 
  summarise(cor = cor(Compressed, Uncompressed, method = "spearman")) |> 
  # Plot heatmap
  ggplot(aes(x = cell_type, y = "Correlation Cell Number per ROI", fill = cor)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(na.value = "gray", limits = c(0, 1), name = "Correlation") + 
  mytheme$standard_new() +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_blank()
  ) +
  ggtitle("Compressed vs Uncompressed") + 
  coord_flip()
print(p)

fn <- file.path(paths$folder_script, paste0(today, "_Correlation_Compressed_Uncompressed", ".png"))
do.call(ggsave, c(list(fn, p), plotsave_param))
```


Finally, calculate the correlation between all cell types per ROI, 
i.e. make a full correlation matrix.

```{r full-correlations-per-roi}
cor_df <- df |> 
  mutate(image_id = stringr::str_remove(image_id, "_Compressed")) |> 
  mutate(image_id = stringr::str_remove(image_id, "_Uncompressed")) |>
  mutate(cell_type = ifelse(cell_type %in% c("T_CD4", "T_CD8"), "T_cell", cell_type)) |>
  dplyr::count(method, case_id, cell_type, image_id) |> # Count occurrences
  tidyr::pivot_wider(names_from = c(cell_type, method), values_from = n, values_fill = 0)

corrplot::corrplot(cor(cor_df[, -c(1:2)]), method = "color", type = "upper", tl.col = "black")
```

Q: What could be other interesting comparisons?